\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage[numbers,square]{natbib}
\usepackage{booktabs}
\usepackage{hyperref}
\hypersetup{hypertexnames=false}
\usepackage{cleveref}

% Theorem environments
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{remark}
\newtheorem{remark}{Remark}[section]
\newtheorem{example}{Example}[section]

\newcommand{\enc}{\mathrm{enc}}
\newcommand{\dec}{\mathrm{dec}}
\newcommand{\fhat}{\hat{f}}
\newcommand{\cipher}[1]{\mathsf{C}(#1)}
\newcommand{\orbitF}{\mathrm{orbit}_F}
\newcommand{\B}{\{0,1\}}

\title{Algebraic Cipher Types: Confidentiality Trade-offs\\in Type Constructors over Trapdoor Computing}
\author{Alexander Towell\\\texttt{lex@metafunctor.com}}
\date{March 2026}

\begin{document}
\maketitle

\begin{abstract}
A cipher map is a total function on bit strings that implements a trapdoor
approximation of a latent function: the untrusted machine evaluates it
blindly; only the trusted machine, holding the secret, can encode inputs
and decode outputs.
We study what happens when algebraic type constructors---products, sums,
and exponentials---are applied to cipher values.
Each constructor forces a specific confidentiality/functionality trade-off
that is absent in plaintext programming.
Our main result is a sum-type impossibility: for $A + B$, either the tag
(which variant) is hidden and pattern matching requires the trapdoor, or
pattern matching is available to the untrusted machine and the tag is
leaked.
No encoding achieves both simultaneously.
We introduce the \emph{orbit closure} of a cipher value under a set of
available operations and prove that confidentiality is bounded by the
fraction of the cipher space reachable from any single value.
A typed-composition discipline turns this bound into a design-time
budget: the type of a cipher program determines how deep its orbit
can go.
We then present two realizations of cipher programs, an expression-tree
decomposition with cipher-map cut points (the practical path, exposed
via a \texttt{@cipher\_node} decorator) and a cipher Turing machine
(space $O(|Q| \cdot |\Gamma|)$, leaks head movement), and evaluate the
cipher Boolean algebra specialization on a 20~Newsgroups Boolean search
task.
\end{abstract}


%% ====================================================================
\section{Introduction}
\label{sec:intro}
%% ====================================================================

Trapdoor computing is a paradigm for evaluating functions on data whose
meaning is hidden behind a one-way trapdoor (a cryptographic hash).
A trusted machine encodes plaintext values into cipher values---opaque bit
strings---and hands them, together with cipher maps (total functions on
bit strings), to an untrusted machine.
The untrusted machine evaluates the cipher maps and returns results.
Without the trapdoor (the secret seed), it cannot decode the results or
determine what function it computed.
The formal framework of cipher maps, their four defining properties (totality,
representation uniformity, correctness, composability), and three concrete
constructions (HashSet, entropy map, trapdoor boolean algebra) are developed
in~\cite{towell2026cipher}.

Ordinary programming languages build complex types from simpler ones
using a small set of type constructors: products ($A \times B$, pairs and
records), sums ($A + B$, tagged unions and variants), and exponentials
($A \to B$, functions).
In plaintext computing, these constructors are transparent: a product
exposes both components, a sum carries a tag that enables pattern matching,
and a function is called by providing an argument.
When values are hidden behind a trapdoor, each of these operations
introduces a confidentiality cost that has no plaintext analogue.

This paper identifies and formalizes those costs.
The main contributions are:

\begin{enumerate}
\item \textbf{Sum-type impossibility} (\Cref{thm:sum-impossibility}).
  For cipher values of sum type $A + B$, tag hiding and pattern matching
  by the untrusted machine are mutually exclusive.
  This is an information-theoretic result, independent of the cipher map
  construction.

\item \textbf{Orbit closure and confidentiality bound}
  (\Cref{sec:orbit}).
  We define the orbit closure of a cipher value under
  available operations, prove monotonicity
  (\Cref{thm:monotonicity}), and derive an upper bound on
  confidentiality in terms of orbit size
  (\Cref{thm:confidentiality-bound}).
  A typed-composition discipline (\Cref{sec:typed-chains}) converts this
  bound into a design-time budget: the type of a cipher program bounds
  its orbit depth and hence its worst-case confidentiality loss.

\item \textbf{Cipher program realizations} (\Cref{sec:realizing}).
  We present two concrete ways to decompose a cipher program into
  cipher-map pieces.
  Expression-tree decomposition with cipher-node annotations is the
  practical path; a cipher Turing machine is a second realization,
  with space $O(|Q| \cdot |\Gamma|)$ at the cost of leaking head
  movement.
  The two realizations share a cut-point structure and differ in what
  residual structure the untrusted machine observes.
\end{enumerate}

We deliberately avoid game-based or simulation-based cryptographic
definitions.
Privacy in trapdoor computing comes from the one-way hash and the
representation uniformity of cipher values, not from access-pattern
indistinguishability.
This is not ORAM~\cite{goldreich1996software}, FHE~\cite{gentry2009fully},
or garbled circuits~\cite{yao1982protocols}; it is a distinct paradigm
where totality and noise closure provide the confidentiality guarantees.


%% ====================================================================
\section{Related Work}
\label{sec:related}
%% ====================================================================

\paragraph{Information flow type systems.}
The sum-type impossibility (\Cref{thm:sum-impossibility}) is closely
related to the implicit flow problem in language-based information
security~\cite{sabelfeld2003language}: a conditional branch on a secret
value leaks information through the control flow path taken.
Our result formalizes this for cipher values: pattern matching on a
cipher sum type is an implicit flow that leaks the tag.
The difference is that information flow type systems prevent the leak
at compile time, while cipher types quantify it at the representation
level.

\paragraph{Functional encryption.}
Functional encryption~\cite{boneh2011functional} allows computing
$f(x)$ on an encrypted $x$ by issuing function-specific decryption
keys.
Cipher maps differ in two ways: (1)~they are information-theoretic
(parameterized by $\eta$, $\varepsilon$, $\delta$) rather than
game-based, and (2)~they are approximate (the untrusted machine
evaluates a total function with bounded error, not an exact
decryption).
The type-constructor trade-offs we identify (tag leakage for sums,
correlation leakage for products) apply equally to functional
encryption schemes that support algebraic operations.

\paragraph{Searchable symmetric encryption.}
SSE originates with Song, Wagner, and
Perrig~\cite{song2000practical} and is placed on rigorous
foundations by Curtmola et al.~\cite{curtmola2006searchable}.
The Boolean search construction in \Cref{sec:cipher-bool} provides
an alternative with information-theoretic parameters: cipher sets
with cipher Boolean AND/OR/NOT, rather than encrypted indices with
leakage profiles.
Subsequent work established that leakage profiles in SSE are
attackable in practice: Islam, Kuzu, and
Kantarcioglu~\cite{islam2012access} demonstrated access-pattern
disclosure attacks; Cash et al.~\cite{cash2015leakage} and
Grubbs, Ristenpart, and Shmatikov~\cite{grubbs2017leakage} mounted
leakage-abuse attacks against deployed SSE and encrypted databases;
Naveed, Kamara, and Wright~\cite{naveed2015inference} did the same for
property-preserving encryption.
Cipher maps avoid property preservation entirely and limit the
adversary to the orbit-closure bound of
\Cref{thm:confidentiality-bound}.

\paragraph{Quantitative information flow.}
The entropy-form confidentiality bound of
\Cref{thm:confidentiality-bound} places this work in the quantitative
information flow tradition~\cite{smith2009foundations}, where leakage
is measured as conditional entropy or min-entropy of the secret given
the adversary's observations.
Our contribution is to specialize this framework to cipher values
under an algebraic operation set, yielding a construction-independent
bound in terms of orbit size.

\paragraph{Related concepts.}
Cipher maps realize the oblivious evaluation of a computable function
(after Turing~\cite{turing1936computable}) through a total function on
bit strings, and the cipher Boolean construction
(\Cref{sec:cipher-bool}) is structurally related to the Bloom
filter~\cite{bloom1970space}, which it generalizes by allocating an
explicit False region alongside the True region.
Our experiments sit on the 20~Newsgroups corpus of
Lang~\cite{lang1995newsweeder}.
Maximizing-confidentiality analyses complementary to this paper's
information-theoretic bound appear in a companion
work~\cite{towell2026maxconf}.

%% ====================================================================
\section{Preliminaries}
\label{sec:prelim}
%% ====================================================================

We recall the cipher map abstraction from~\cite{towell2026cipher}.
All notation follows that paper; we cite rather than re-derive.

\begin{definition}[Cipher map~{\cite[Def.~1.1]{towell2026cipher}}]
\label{def:cipher-map}
A \emph{cipher map} for a latent function $f : X \to Y$ is a tuple
$(\fhat, \enc, \dec, s)$ where
$\fhat : \B^n \to \B^n$ is a total function on $n$-bit strings,
$\enc : X \times \{0,\ldots,K{-}1\} \to \B^n$ is an encoding function,
$\dec : \B^n \to Y \cup \{\bot\}$ is a decoding function, and
$s$ is a secret (the trapdoor).
The untrusted machine holds $\fhat$.
The trusted machine holds $\enc$, $\dec$, and $s$.
\end{definition}

A cipher map satisfies four properties, parameterized by
$(\eta, \varepsilon, \mu, \delta)$~\cite[Sec.~4]{towell2026cipher}:

\begin{enumerate}
\item \textbf{Totality.} $\fhat$ is defined on all $2^n$ inputs.
  Out-of-domain outputs are indistinguishable from uniform under the
  random oracle model~\cite{bellare1993random}.

\item \textbf{Representation uniformity} ($\delta$-bounded).
  The marginal distribution of cipher values is $\delta$-close (in total
  variation distance) to uniform over $\B^n$.

\item \textbf{Correctness} ($\eta$-bounded).
  For a random in-domain element $x$ and random representation index $k$,
  $\Pr[\dec(\fhat(\enc(x,k))) \neq f(x)] \leq \eta$.

\item \textbf{Composability.}
  For cipher maps $\fhat$ (correctness $\eta_f$) and $\hat{g}$ (correctness
  $\eta_g$), the composition $\hat{g} \circ \fhat$ has correctness
  $\eta_{g \circ f} = 1 - (1-\eta_f)(1-\eta_g)$.
\end{enumerate}

The Bernoulli error model~\cite{towell2026bernoulli} provides the
quantitative foundation: element-wise independence of errors reduces the
composition analysis from exponentially many joint failure modes to
per-element Bernoulli parameters.
We use $h : \B^* \to \B^n$ throughout to denote a cryptographic hash
modeled as a random oracle.

A cipher value $\enc(v, k)$ is a cipher map for the constant function
$f_v(x) = v$; we use ``cipher value'' and ``cipher map from a trivial
domain'' interchangeably.

\begin{definition}[Trusted and untrusted machines~{\cite[Sec.~5]{towell2026cipher}}]
\label{def:machines}
The \emph{trusted machine} $T$ holds the trapdoor $s$, $\enc$, $\dec$, and
can encode, decode, and inject filler queries.
The \emph{untrusted machine} $U$ holds $\fhat$ and a collection of cipher
values.
$U$ can evaluate $\fhat$ on any bit string; it cannot decode, distinguish
real from filler queries, or determine the domain $X$.
\end{definition}


%% ====================================================================
\section{Cipher Type Constructors}
\label{sec:types}
%% ====================================================================

We now examine how the standard algebraic type constructors interact with
cipher values.
For each constructor, there is a choice: apply the cipher operation to the
entire constructed type, or construct from already-ciphered components.
These two strategies yield different confidentiality and functionality
properties.
We write $\cipher{X}$ for the type of cipher values encoding elements of
$X$.


%% --------------------------------------------------------------------
\subsection{Void and Unit}
\label{sec:void-unit}
%% --------------------------------------------------------------------

The void type $\mathbf{0}$ has no values.
A cipher map $\fhat$ for a function with domain $\mathbf{0}$ is vacuously
correct ($\eta = 0$) and carries no information---it is a total function
on $\B^n$ that is never queried on a valid encoding.
From the untrusted machine's perspective, $\fhat$ is indistinguishable from
a random function.

The unit type $\mathbf{1}$ has exactly one value, $\star$.
A cipher value $\enc(\star, k) \in \B^n$ encodes $\star$; with $K(\star) \geq 1$
representations, all cipher values are representations of the same
underlying value.
The untrusted machine knows that every valid cipher value encodes $\star$
but cannot distinguish valid encodings from noise (by totality).
A cipher map $\fhat : \B^n \to \B^n$ for a function from $\mathbf{1}$ is
a cipher constant: it maps every valid encoding of $\star$ to a cipher
encoding of $f(\star)$.


%% --------------------------------------------------------------------
\subsection{Product Types}
\label{sec:products}
%% --------------------------------------------------------------------

The product type $A \times B$ consists of pairs $(a, b)$ with $a \in A$ and
$b \in B$.
In plaintext computing, a product comes with two projections
$\pi_1 : A \times B \to A$ and $\pi_2 : A \times B \to B$.
Under cipher encoding, we have two choices.

\paragraph{Joint encoding: $\cipher{A \times B}$.}
Encode the pair $(a, b)$ as a single cipher value $\enc((a,b), k) \in \B^n$.
The untrusted machine sees one opaque bit string.
It cannot determine $a$ or $b$ individually, nor can it detect correlations
between $a$ and $b$.
But the projections $\pi_1, \pi_2$ are not available: to extract $a$ from a
cipher encoding of $(a,b)$, one would need a cipher map
$\hat{\pi}_1 : \B^n \to \B^n$ such that
$\dec_A(\hat{\pi}_1(\enc((a,b),k))) = a$---which requires a separate cipher
map construction for each projection.

\paragraph{Component-wise encoding:
$\cipher{A} \times \cipher{B}$.}
Encode $a$ and $b$ independently as
$(\enc_A(a, k_1),\allowbreak \enc_B(b, k_2))$.
The untrusted machine sees a pair of cipher values.
Projections are trivially available (just take the first or second
component).
But the joint distribution of $(a, b)$ is no longer hidden: even if each
component has marginal representation uniformity ($\delta \approx 0$),
the correlation between $a$ and $b$ is preserved in the joint distribution
of the cipher pair.
This is the encoding granularity principle
of~\cite[Sec.~9]{towell2026cipher}: coarser encoding hides more
correlations but blocks more operations.

\begin{proposition}[Product confidentiality trade-off]
\label{prop:product-tradeoff}
Let $A$ and $B$ be finite types.
\begin{enumerate}
\item Under joint encoding $\cipher{A \times B}$, the untrusted machine
  cannot distinguish $(a_1, b_1)$ from $(a_2, b_2)$ (up to $\delta$),
  but projections require a cipher map (and thus a separate construction
  by the trusted machine).
\item Under component-wise encoding $\cipher{A} \times \cipher{B}$,
  projections are free, but an adversary observing $N$ pairs
  $(\enc_A(a_i, k_{1i}), \enc_B(b_i, k_{2i}))$ can estimate the joint
  distribution of $(A, B)$ to arbitrary precision as $N \to \infty$.
\end{enumerate}
\end{proposition}

\begin{proof}
Part~(1): Under joint encoding with representation uniformity $\delta$,
each cipher value is $\delta$-close to uniform over $\B^n$.
Without the trapdoor, the untrusted machine cannot decode.
Projections require mapping a cipher encoding of $(a,b)$ to a cipher
encoding of $a$; this is a function from the code space of $A \times B$ to
the code space of $A$, which is exactly a cipher map and requires
construction by the trusted machine.

Part~(2): Under component-wise encoding, the $i$-th observation is
$(\enc_A(a_i, k_{1i}),\allowbreak \enc_B(b_i, k_{2i}))$.
Even with perfect marginal uniformity ($\delta = 0$), each $\enc_A$ is a
deterministic function of $(a_i, k_{1i})$.
If the same representation index is reused, repeated values produce
identical cipher values; even without reuse, the number of distinct cipher
values associated with each latent value is finite ($K(a)$ for each $a$).
Given $N$ observations, the adversary builds the empirical joint distribution
over cipher-value pairs.
Since each cipher value maps to at most one latent value (given a fixed
representation index), the empirical joint distribution of cipher pairs
converges to a mixture of the latent joint distribution.
By standard convergence results, $N = O(|A| \cdot |B| / \xi^2)$
observations suffice to estimate the joint to accuracy $\xi$ in total
variation distance.
\end{proof}


%% --------------------------------------------------------------------
\subsection{Sum Types}
\label{sec:sums}
%% --------------------------------------------------------------------

The sum type $A + B$ is a tagged union: each value is either
$\mathrm{inl}(a)$ for some $a \in A$, or $\mathrm{inr}(b)$ for some
$b \in B$.
The tag (left vs.\ right) enables pattern matching: given a value of type
$A + B$, one inspects the tag to determine which handler to apply.

In plaintext computing, the tag is always available and pattern matching
is free.
Under cipher encoding, this is no longer the case.

\begin{theorem}[Sum-type impossibility]
\label{thm:sum-impossibility}
Let $A$ and $B$ be non-empty finite types with $|A| \geq 1$ and
$|B| \geq 1$.
For cipher values of sum type $A + B$, the following two properties are
mutually exclusive:
\begin{enumerate}
\item \textbf{Tag hiding}: the untrusted machine cannot determine, from a
  cipher value alone, whether it encodes an $A$-value or a $B$-value with
  advantage better than $\delta$ (the representation uniformity parameter).
\item \textbf{Untrusted pattern matching}: the untrusted machine can route
  a cipher value to the correct handler ($A$-handler or $B$-handler)
  without assistance from the trusted machine.
\end{enumerate}
\end{theorem}

\begin{proof}
We prove the two directions by exhibiting the information-theoretic
constraint.

\medskip
\noindent\textbf{Joint encoding ($\cipher{A+B}$) achieves tag hiding but
blocks pattern matching.}
Under joint encoding, each value $v \in A + B$ (whether $\mathrm{inl}(a)$ or
$\mathrm{inr}(b)$) is encoded as a single cipher value $\enc(v, k) \in \B^n$.
Representation uniformity ($\delta$-bounded) guarantees that the marginal
distribution of $\enc(v, k)$ over random $v$ and $k$ is $\delta$-close to
uniform on $\B^n$.
In particular, the set of cipher values encoding $A$-values and the set
encoding $B$-values are not efficiently distinguishable by the untrusted
machine: both are subsets of $\B^n$, and their union is $\delta$-close to
covering $\B^n$ uniformly.
Therefore the tag (which variant) is hidden up to $\delta$.

Now suppose the untrusted machine could nonetheless perform pattern matching:
given $c = \enc(v, k)$, it could determine whether to apply the $A$-handler
or the $B$-handler.
This requires a function $\tau : \B^n \to \{A, B\}$ such that
$\tau(\enc(\mathrm{inl}(a), k)) = A$ for all $a, k$ and
$\tau(\enc(\mathrm{inr}(b), k)) = B$ for all $b, k$.
But $\tau$ is exactly a distinguisher between $A$-encodings and
$B$-encodings.
Such a distinguisher exists only if the two sets of cipher values are
disjoint (or nearly so) in a way detectable without the trapdoor.
If $\tau$ succeeds with probability $1 - \gamma$, then the tag is leaked
with advantage at least $1 - \gamma - \delta$, contradicting tag hiding
when $\gamma$ is small.

\medskip
\noindent\textbf{Component-wise encoding ($\cipher{A} + \cipher{B}$)
permits pattern matching but leaks the tag.}
Under component-wise encoding, the value is stored as a tagged pair:
either $(\texttt{L}, \enc_A(a, k))$ or $(\texttt{R}, \enc_B(b, k))$,
where the tag $\texttt{L}$ or $\texttt{R}$ is in the clear (or
distinguishable by the encoding structure---for instance, $A$-cipher-values
and $B$-cipher-values occupy disjoint regions of bit-string space, or
carry a type identifier).
The untrusted machine reads the tag and routes to the correct handler.
Pattern matching is available.

But the tag is observable.
Given a cipher value, the untrusted machine learns whether it encodes an
$A$-value or a $B$-value.
For any distribution $D$ on $A + B$, the fraction of queries that are
$A$-values is observable, leaking $\Pr_{v \sim D}[v \in A]$.

\medskip
\noindent\textbf{No intermediate encoding achieves both.}
Suppose an encoding achieves both tag hiding (advantage $\leq \delta'$
for some small $\delta'$) and untrusted pattern matching (success
probability $\geq 1 - \gamma$).
Pattern matching requires a function $\tau : \B^n \to \{A, B\}$ that is
correct on valid encodings with probability $\geq 1 - \gamma$.
Tag hiding requires that no such function exists with advantage better than
$\delta'$.
Let $p = \Pr[v \in A]$ be the prior probability of an $A$-value.
A trivial strategy that ignores the encoding and always guesses the
majority type achieves accuracy $\max(p, 1-p)$.
The advantage of $\tau$ over this baseline is at least
$(1 - \gamma) - \max(p, 1-p)$.
Tag hiding requires this advantage to be at most $\delta'$.
Therefore $(1-\gamma) - \max(p, 1-p) \leq \delta'$, which means either
$\gamma \geq 1 - \max(p, 1-p) - \delta'$ (pattern matching fails often)
or $\delta' \geq (1-\gamma) - \max(p, 1-p)$ (the tag is substantially
leaked).
For the balanced case $p = 1/2$, this simplifies to
$1/2 - \gamma \leq \delta'$.
Unbalanced distributions ($p \neq 1/2$) make the impossibility stronger:
the baseline accuracy is higher, leaving less room for the encoding to
add information without leaking the tag.
\end{proof}

\begin{remark}[Contrast with products]
\label{rem:sum-vs-product}
Products leak correlations; sums leak the tag.
The asymmetry is structural.
For a product $A \times B$, the two components are always both present;
the question is whether their relationship (correlation) is visible.
For a sum $A + B$, only one component is present; the question is whether
which one is visible.
A product hides \emph{what values co-occur}; a sum hides \emph{what type
is occupied}.
\end{remark}

\begin{remark}[Trusted pattern matching]
\label{rem:trusted-matching}
Under joint encoding $\cipher{A+B}$, pattern matching is still possible---by
the trusted machine.
The trusted machine decodes $\dec(c)$, inspects the tag, applies the
appropriate handler in plaintext, and re-encodes the result.
This requires a round trip through the trusted machine for every case
analysis.
The cost is a communication round, not an impossibility.
The impossibility is for \emph{untrusted} pattern matching without the
trapdoor.
\end{remark}

\begin{example}[Cipher optionals]
\label{ex:optional}
The optional type $\mathrm{Maybe}(A) = \mathbf{1} + A$ is a sum.
Under $\cipher{\mathbf{1} + A}$, the untrusted machine cannot tell
whether a cipher value encodes $\mathrm{Nothing}$ or $\mathrm{Just}(a)$.
Under $\cipher{\mathbf{1}} + \cipher{A}$, the untrusted machine knows
whether the value is present, but not what it is.
For a database application, the first encoding hides which records have null
fields; the second reveals the null pattern but permits null-checking without
the trapdoor.
\end{example}


%% --------------------------------------------------------------------
\subsection{Exponential Types}
\label{sec:exponentials}
%% --------------------------------------------------------------------

The exponential type $B^A = A \to B$ is the type of functions from $A$ to
$B$.
A cipher map $\fhat$ for $f : A \to B$ is already an element of the cipher
exponential type: it is a total function on $\B^n$ that, when restricted
to valid encodings of $A$, produces (approximate) encodings of $B$-values.

This observation, developed in~\cite{towell2026cipher}, means the cipher
exponential type is not a new construction but the cipher map abstraction
itself.
The four properties (totality, representation uniformity, correctness,
composability) are exactly the conditions that make $\fhat$ a well-behaved
element of $\cipher{A \to B}$.

The encoding granularity trade-off from \Cref{sec:products} extends to
exponentials.
Given cipher maps $\fhat : \B^n \to \B^n$ and $\hat{g} : \B^n \to \B^n$,
applying them independently (component-wise) to the two components of a
product cipher value reveals that both functions are being applied---leaking
the computation pattern.
Encoding the pair of function applications as a single cipher map
$\widehat{(f,g)} : \B^n \to \B^n$ for the product function
$(f, g)(a, b) = (f(a), g(b))$ hides the computation pattern but requires
a dedicated construction.


%% ====================================================================
\section{Orbit Closure and Information Leakage}
\label{sec:orbit}
%% ====================================================================

The type constructors of \Cref{sec:types} describe static confidentiality
properties: what the untrusted machine can learn from the structure of a
single cipher value.
We now develop a tool for analyzing dynamic confidentiality: what the
untrusted machine can learn by actively applying the cipher maps it holds.


%% --------------------------------------------------------------------
\subsection{Definition}
\label{sec:orbit-def}
%% --------------------------------------------------------------------

\begin{definition}[Orbit closure]
\label{def:orbit}
Let $F = \{\fhat_1, \ldots, \fhat_k\}$ be a set of cipher maps
(total functions $\B^n \to \B^n$) available to the untrusted machine,
and let $c \in \B^n$ be a cipher value.
The \emph{orbit closure} of $c$ under $F$ is the smallest set
$\orbitF(c) \subseteq \B^n$ such that:
\begin{enumerate}
\item $c \in \orbitF(c)$, and
\item if $c' \in \orbitF(c)$ and $\fhat_i \in F$, then
  $\fhat_i(c') \in \orbitF(c)$.
\end{enumerate}
Equivalently, $\orbitF(c)$ is the set of all bit strings reachable from
$c$ by composing operations in $F$ in any order and any number of times.
\end{definition}

Since $\B^n$ is finite, $\orbitF(c)$ is always finite and can be computed
by iterative expansion (breadth-first search from $c$, applying each
$\fhat_i$ at each step, until no new values appear).

\begin{remark}
The orbit closure depends only on the cipher maps $F$ available to the
untrusted machine and the starting cipher value $c$.
It does not depend on the latent function, the domain, or the trapdoor.
This makes it a tool for analysis by the trusted machine designer: given
the operations you plan to expose, how much can the untrusted machine
explore?
\end{remark}


%% --------------------------------------------------------------------
\subsection{Monotonicity}
\label{sec:monotonicity}
%% --------------------------------------------------------------------

\begin{theorem}[Monotonicity of orbit closure]
\label{thm:monotonicity}
If $F \subseteq G$ (every operation in $F$ is also in $G$), then
$\orbitF(c) \subseteq \mathrm{orbit}_G(c)$ for all $c \in \B^n$.
\end{theorem}

\begin{proof}
By induction on the iterative construction of $\orbitF(c)$.

\emph{Base case.} $c \in \orbitF(c)$ and $c \in \mathrm{orbit}_G(c)$
by definition.

\emph{Inductive step.} Suppose $c' \in \orbitF(c)$ and $c' \in
\mathrm{orbit}_G(c)$ (inductive hypothesis).
For any $\fhat_i \in F$, we have $\fhat_i(c') \in \orbitF(c)$ by
\Cref{def:orbit}.
Since $F \subseteq G$, $\fhat_i \in G$, so
$\fhat_i(c') \in \mathrm{orbit}_G(c)$ by \Cref{def:orbit} applied to $G$.

Every element of $\orbitF(c)$ is therefore also in $\mathrm{orbit}_G(c)$.
\end{proof}

The interpretation is direct: exposing more operations to the untrusted
machine can only enlarge (never shrink) the set of cipher values it can
reach.
Each additional operation is an additional edge in the reachability graph
over $\B^n$.

\begin{corollary}
\label{cor:empty-ops}
$\mathrm{orbit}_\emptyset(c) = \{c\}$ for all $c$.
If the untrusted machine has no cipher maps, it learns nothing beyond the
cipher value it was given.
\end{corollary}

\begin{proof}
With $F = \emptyset$, only the base case of the orbit definition applies.
\end{proof}


%% --------------------------------------------------------------------
\subsection{Confidentiality Bound}
\label{sec:conf-bound}
%% --------------------------------------------------------------------

The orbit closure quantifies reachability.
We now connect it to confidentiality via an information-theoretic
bound: each distinct cipher value in $\orbitF(c)$ is a signal the
adversary observes, and the number of such signals upper-bounds how
much can be learned about the latent value.

Let $X$ denote the random latent value encoded by $c$ (drawn from the
distribution over inputs that the trusted machine uses) and let
$\mathcal{V}_F(c) = \orbitF(c)$ denote the adversary's view: the set
of cipher values reachable from $c$ under $F$.
The adversary's residual uncertainty about $X$ is quantified by the
conditional entropy $H(X \mid \mathcal{V}_F(c))$.

\begin{theorem}[Confidentiality bound, entropy form]
\label{thm:confidentiality-bound}
Let $F$ be a set of cipher maps available to the untrusted machine and
$c$ a cipher value.
Then
\[
H(X \mid \mathcal{V}_F(c)) \;\geq\; H(X) - \log_2 |\orbitF(c)|.
\]
\end{theorem}

\begin{proof}
The view $\mathcal{V}_F(c)$ takes at most $|\orbitF(c)|$ distinct
values as $X$ ranges over its distribution, since the orbit is a
finite set of bit strings computable from $c$.
Therefore
\[
I(X; \mathcal{V}_F(c)) \;\leq\; H(\mathcal{V}_F(c)) \;\leq\; \log_2 |\orbitF(c)|,
\]
and $H(X \mid \mathcal{V}_F(c)) = H(X) - I(X; \mathcal{V}_F(c))$.
\end{proof}

\begin{corollary}[Confidentiality bound, set form]
\label{cor:confidentiality-set}
Let $S_F(c) \subseteq X$ denote the set of latent values the adversary
cannot rule out from $\mathcal{V}_F(c)$, and define the
\emph{confidentiality} of $c$ under $F$ as
$\mathrm{conf}_F(c) = 1 - |S_F(c)|/|X|$.
Then
\[
\mathrm{conf}_F(c) \;\geq\; 1 - \frac{|\orbitF(c)|}{|X|}.
\]
\end{corollary}

\begin{proof}
Each element of $S_F(c)$ is distinguished from the others by at least
one observable feature of the view, and the view takes at most
$|\orbitF(c)|$ distinct values, so
$|S_F(c)| \leq |\orbitF(c)|$.
The bound follows.
\end{proof}

\begin{remark}[Loose bound, tight in the worst case]
Both forms are loose: they ignore the functional relationships between
orbit elements (the adversary knows $c' = \fhat(c)$, which constrains
the possible latent values beyond mere set size) and the algebraic
structure of $F$.
Tighter bounds for specific operations are an open problem.
Equality is approached when orbit elements behave like independent
signals about $X$, e.g., when the cipher maps in $F$ are modeled as
random oracles~\cite{bellare1993random}.
\end{remark}

\begin{remark}[Extremes]
When $|\orbitF(c)| \geq |X|$, the set-form bound is vacuous; in
principle the orbit can identify every latent value.
When $|\orbitF(c)| = 1$ (no operations available, per
\Cref{cor:empty-ops}), confidentiality is maximal:
$H(X \mid \mathcal{V}_F(c)) = H(X)$ and $\mathrm{conf}_F(c) = 1 - 1/|X|$.
\end{remark}


%% --------------------------------------------------------------------
\subsection{Examples}
\label{sec:orbit-examples}
%% --------------------------------------------------------------------

\begin{example}[Boolean operations]
\label{ex:boolean-orbit}
Let $X = \{0, 1\}$ (Boolean) with cipher maps $\widehat{\mathrm{AND}}$ and
$\widehat{\mathrm{NOT}}$ available to the untrusted machine.
Given a cipher value $c$ encoding some unknown Boolean $x$, the untrusted
machine can compute:
\begin{align*}
c_1 &= \widehat{\mathrm{NOT}}(c), \\
c_2 &= \widehat{\mathrm{AND}}(c, c_1)
     = \widehat{\mathrm{AND}}(c, \widehat{\mathrm{NOT}}(c)).
\end{align*}

Now, $\mathrm{AND}(x, \mathrm{NOT}(x)) = 0$ for all $x$ (up to
correctness $\eta$).
So $c_2$ is an (approximate) cipher encoding of the constant
$\mathrm{FALSE}$.
If the untrusted machine has a reference cipher value $c_{\mathrm{F}}$
known to encode $\mathrm{FALSE}$ (for instance, from a prior computation
$\widehat{\mathrm{AND}}(c', \widehat{\mathrm{NOT}}(c'))$ for some other
cipher value $c'$), it can compare $c_2$ with $c_{\mathrm{F}}$.

This comparison reveals information: if $c_2 = c_{\mathrm{F}}$
(same representation), the untrusted machine confirms the tautology
and potentially identifies the representation index.
Writing $F_0 = \{\widehat{\mathrm{AND}}, \widehat{\mathrm{NOT}}\}$, the
orbit $\mathrm{orbit}_{F_0}(c)$ includes $c$, $c_1$, $c_2$, and all further
compositions---quickly covering a nontrivial fraction of the cipher space
for small $n$.
\end{example}

\begin{example}[Successor on a bounded type]
\label{ex:successor-orbit}
Let $X = \{0, 1, \ldots, m-1\}$ (integers mod $m$) with a single cipher
map $\widehat{\mathrm{succ}}$ implementing the successor function
$x \mapsto (x+1) \bmod m$.
Given any cipher value $c$ encoding some $x$:
\[
\mathrm{orbit}_{\{\widehat{\mathrm{succ}}\}}(c) =
  \{c,\; \widehat{\mathrm{succ}}(c),\;
   \widehat{\mathrm{succ}}^2(c),\; \ldots,\;
   \widehat{\mathrm{succ}}^{m-1}(c)\}.
\]
After $m$ applications, the orbit returns to $c$ (since successor is
cyclic on $\mathbb{Z}/m\mathbb{Z}$).
The orbit has exactly $m$ elements (assuming $\eta = 0$).
The untrusted machine does not learn which element is
$\enc(0, k)$, but it learns the cycle structure: there exists a cycle of
length $m$ starting from $c$.
This reveals $m = |X|$, the domain cardinality.

By \Cref{thm:confidentiality-bound}, the confidentiality is
$1 - m/2^n$.
For $m \ll 2^n$, this is close to 1: the orbit covers a tiny fraction
of the cipher space.
For $m$ close to $2^n$, confidentiality degrades.
\end{example}

\begin{example}[Branching]
\label{ex:branching-orbit}
Suppose the untrusted machine holds cipher maps $\fhat$ and $\hat{g}$
and evaluates both on a cipher value $c$, obtaining $\fhat(c)$ and
$\hat{g}(c)$.
The orbit of $c$ under $\{\fhat, \hat{g}\}$ includes both results plus all
further compositions.
If $\fhat$ and $\hat{g}$ are the two branches of a conditional
(the ``then'' and ``else'' handlers), both cipher results are visible to
the untrusted machine---even though in plaintext computing, only one
branch would be taken.

This is a form of information leakage specific to trapdoor computing:
the untrusted machine does not know which branch is ``correct,'' but it
sees both outcomes, enlarging the orbit and reducing confidentiality.
Mitigating this requires either (a)~evaluating both branches and having the
trusted machine select the correct result, or (b)~encoding the branch
selection as part of the cipher map (a joint encoding of the conditional).
Strategy (a) doubles the computation; strategy (b) requires a dedicated
cipher map construction.
\end{example}


\begin{remark}[Active probing via Boolean operations]
\label{rem:active-probing}
Exposing cipher Boolean operations creates an active probing risk
beyond passive observation.
Given cipher AND and a cipher value $c$, the adversary computes
$\mathrm{AND}(c, c)$, $\mathrm{AND}(c, c')$ for other values $c'$,
and checks structural consistency.
Over many probes, the adversary partitions cipher values into
equivalence classes corresponding to latent Boolean values.
Each probe enlarges the orbit, and confidentiality degrades per
\Cref{thm:confidentiality-bound}.
This is a concrete instance of orbit closure: Boolean operations are
probing tools that trade functionality for confidentiality.
\end{remark}

\subsection{Typed Composition Chains}
\label{sec:typed-chains}

The active probing risk motivates controlling the orbit by
construction.
The orbit grows because the adversary can feed cipher map outputs back
as inputs, composing operations to reach new cipher values.
A direct mitigation: \emph{prevent self-composition by typing the
cipher spaces}.

\begin{definition}[Typed composition chain]
\label{def:typed-chain}
A \emph{typed composition chain} of depth $k$ is a sequence of
cipher maps $\fhat_0, \ldots, \fhat_{k-1}$ where
$\fhat_i : \cipher{A}_i^{a_i} \to \cipher{A}_{i+1}$ has arity
$a_i \geq 1$ over the cipher space $\cipher{A}_i$, and the cipher
spaces $\cipher{A}_0, \ldots, \cipher{A}_k$ are \emph{distinct}
(disjoint hash value sets, different secrets).
\end{definition}

Since $\cipher{A}_i \neq \cipher{A}_j$ for $i \neq j$, the output of
$\fhat_i$ cannot be fed into $\fhat_j$ for $j \neq i+1$.
The adversary can compose at most $k$ operations in sequence and cannot
loop.

\begin{proposition}[Orbit bound for typed chains]
\label{prop:typed-orbit}
Let $V_0 \subseteq \cipher{A}_0$ be a set of $m$ initial cipher values
available to the adversary, and let $F = \{\fhat_0, \ldots, \fhat_{k-1}\}$
be a typed composition chain of depth $k$ with arities
$a_0, \ldots, a_{k-1}$.
Define $N_0 = m$ and $N_{i+1} = N_i^{a_i}$.
Then
\[
|\orbitF(V_0)| \;\leq\; \sum_{i=0}^{k} N_i,
\]
a finite bound determined entirely by $(k, m, a_0, \ldots, a_{k-1})$
and independent of the cipher-space sizes $|\cipher{A}_i|$.
Two useful special cases:
\begin{itemize}
\item \textbf{Single-value start} ($m = 1$), any arities:
  $|\orbitF(V_0)| \leq 1 + k$.
\item \textbf{Unary chain} ($a_i = 1$), any $m$:
  $|\orbitF(V_0)| \leq m(k+1)$.
\end{itemize}
The corresponding confidentiality bound follows from
\Cref{cor:confidentiality-set}.
\end{proposition}

\begin{proof}
Partition $\orbitF(V_0)$ by level: $L_0 = V_0 \subseteq \cipher{A}_0$,
and $L_{i+1} \subseteq \cipher{A}_{i+1}$ is the image of $\fhat_i$
applied to all $a_i$-tuples of elements of $L_i$.
Distinct cipher spaces prevent back-references, so every orbit element
lies in exactly one $L_i$.
Since $|L_i| \leq |L_{i-1}|^{a_{i-1}}$, induction gives $|L_i| \leq N_i$,
and therefore $|\orbitF(V_0)| = \sum_i |L_i| \leq \sum_i N_i$.
For $m = 1$, $N_i = 1$ at every level (identical inputs produce one
output for a deterministic cipher map), so the sum is $1+k$.
For $a_i = 1$, $N_i = m$ at every level, so the sum is $m(k+1)$.
\end{proof}

\begin{example}[Depth-limited Boolean search]
\label{ex:typed-bool}
Consider Boolean search with maximum query depth 2 (e.g., $w_1 \land w_2$
but not $w_1 \land w_2 \land w_3$):
\begin{itemize}
\item Cipher sets output to $\cipher{\mathrm{Bool}}_0$.
\item $\mathrm{AND}_0 : \cipher{\mathrm{Bool}}_0 \times
  \cipher{\mathrm{Bool}}_0 \to \cipher{\mathrm{Bool}}_1$
  (binary, $a_0 = 2$).
\item $\mathrm{AND}_1 : \cipher{\mathrm{Bool}}_1 \times
  \cipher{\mathrm{Bool}}_1 \to \cipher{\mathrm{Bool}}_2$
  (binary, $a_1 = 2$).
\item No $\mathrm{AND}_2$ is defined.
\end{itemize}
Starting from a single cipher Boolean $c$ (one membership-query
output), the orbit has size at most $1 + k = 3$: namely
$\{c, \mathrm{AND}_0(c, c), \mathrm{AND}_1(\mathrm{AND}_0(c, c),
\mathrm{AND}_0(c, c))\}$.
With $m$ membership-query outputs as initial values, the bound is
$m + m^2 + m^4$, still finite and controlled by the chain depth.
Contrast with a single shared cipher Boolean space, where the
adversary can self-compose indefinitely and the orbit grows up to
$2^n$.
\end{example}

The general principle: \textbf{the type system controls the orbit}.
By choosing how many cipher space levels to define, the system designer
sets the maximum composition depth and therefore the maximum orbit size
and minimum confidentiality.
This is a compile-time (construction-time) decision with no runtime
cost: the untrusted machine simply does not have cipher maps for
deeper levels.


%% ====================================================================
\section{Realizing Cipher Programs}
\label{sec:realizing}
%% ====================================================================

The type-theoretic results of \Cref{sec:types,sec:orbit} bound what any
cipher-program realization can achieve.
We now examine two concrete realizations, each decomposing a program
into cipher-map pieces along a different axis.
A cipher Turing machine treats the program as a repeated transition
step; expression-tree decomposition treats the program as a DAG whose
selected subtrees are replaced by cipher maps.
The two realizations leak different residual structure and occupy
different points in the space-time-confidentiality trade-off.
We present them in turn and then make explicit the common cut-point
pattern they share (\Cref{sec:tm-vs-tree}).


%% --------------------------------------------------------------------
\subsection{Cipher Turing Machines}
\label{sec:cipher-tm}
%% --------------------------------------------------------------------

A cipher map for $f : X \to Y$ is a lookup table of size $O(|X|)$, which
is prohibitive when $|X|$ is large (e.g., string processing with
$|X| = |\Sigma|^\ell$).
A Turing machine is an alternative whose space depends on the
\emph{program}, not the \emph{data}.

\begin{definition}[Cipher Turing machine]
\label{def:cipher-tm}
A \emph{cipher Turing machine} consists of a finite state set $Q$, a
tape alphabet $\Gamma$ with blank $\sqcup$, an initial state $q_0$, a
halting state $q_\mathrm{halt}$, and a cipher map $\hat{\delta}_T$ for
the transition function
$\delta_T : Q \times \Gamma \to Q \times \Gamma \times \{L, R, S\}$.
Each tape cell is a single cipher value in $\cipher{Q \times \Gamma}$:
the cell stores a cipher encoding of the (state, symbol) pair at that
cell, with the distinguished state $\varnothing \in Q$ marking cells
without the head.
On entry, the trusted machine encodes the head position by placing a
cipher value for $(q_0, \sqcup)$ (or the first input symbol) at the
start cell and cipher encodings of $(\varnothing, \sqcup)$ elsewhere.
At each step the untrusted machine applies $\hat{\delta}_T$ to the
current-head cell, obtaining a pair $(c', d)$ where $c'$ is a cipher
value in $\cipher{Q \times \Gamma}$ and $d \in \{L, R, S\}$ is the head
direction in the clear.
The untrusted machine writes $c'$ back to the current cell, shifts the
``head'' marker according to $d$ (by swapping the state component of
the neighbouring cell with the current cell via a second cipher map
for the shift operation), and repeats.
The head direction is not cipher-encoded: the untrusted machine must
know which cell to access next.
Ciphering the direction would require evaluating all three neighbouring
cells on every step, tripling per-step work while still revealing that
a step occurred.
\end{definition}

\begin{proposition}[Cipher TM space complexity]
\label{prop:cipher-tm-space}
The cipher map $\hat{\delta}_T$ has space $O(|Q| \cdot |\Gamma|)$ (plus
$-\log_2 \varepsilon$ bits per entry for the noise budget and
$H(Q \times \Gamma \times \{L,R,S\})$ bits per entry for value encoding,
per~\cite[Sec.~6]{towell2026cipher}), independent of the input domain
size $|X|$.
When $|X| \gg |Q| \cdot |\Gamma|$, this is a qualitative improvement
over the $O(|X|)$ space of a lookup-table cipher map, at the cost of
$O(T(|x|))$ steps per evaluation instead of $O(1)$, where $T$ is the
running time of the plaintext TM.
\end{proposition}

\begin{proposition}[Head-movement leakage]
\label{prop:head-determinism}
For a deterministic cipher TM with $\eta = 0$, the head-movement
sequence $d_1, d_2, \ldots \in \{L, R, S\}^*$ is a deterministic
function of the input.
Two inputs that produce different head-movement sequences are
distinguishable by the untrusted machine without the trapdoor.
\end{proposition}

\begin{proof}
With $\eta = 0$, $\hat{\delta}_T$ reproduces the plaintext Turing
machine's transitions exactly, so the head-movement sequence is
identical to the plaintext sequence, which is a deterministic function
of the input.
Distinct sequences are directly observable by the untrusted machine.
\end{proof}

\begin{remark}[Comparison with ORAM]
\label{rem:oram}
Oblivious RAM~\cite{goldreich1996software} is designed precisely to
hide access patterns.
A cipher TM makes no such guarantee: tape contents and the transition
function are opaque, but the sequence of cells visited is not.
This is weaker than ORAM but requires no $O(\log N)$ per-access
overhead.
Non-zero $\eta$ injects noise into the head-movement sequence, but the
noise is seed-systematic rather than random, so it is unreliable as an
access-pattern obfuscation mechanism.
\end{remark}


%% --------------------------------------------------------------------
\subsection{Expression-Tree Decomposition}
\label{sec:cipher-programs}
%% --------------------------------------------------------------------

The type-theoretic trade-offs of \Cref{sec:products,sec:sums} have a
direct operational realization: given a program represented as an
expression tree, which subtrees should be replaced by cipher maps?

\begin{definition}[Cipher node annotation]
\label{def:cipher-node}
Given a program $P$ represented as an expression tree, a
\emph{cipher node annotation} marks a subset of nodes $\{n_1, \ldots,
n_k\}$ in the tree.  Each marked node $n_i$, together with its entire
subtree, is replaced by a cipher map: the subtree is evaluated at
construction time (on the trusted machine) over a specified domain, and
the resulting mapping is stored as a cipher map backed by a
perfect hash function~\cite{fredman1984storing}.
Unmarked nodes above the cipher nodes become a \emph{combiner}: itself
a cipher map built over the hash-value outputs of the components.
\end{definition}

The placement of cipher node annotations controls the encoding
granularity:

\begin{itemize}
\item \textbf{Annotation at the root}: the entire program is one cipher
  map.  Maximum confidentiality (no intermediate values exposed), maximum
  space ($O(|X|)$ where $X$ is the input domain of the full program).
\item \textbf{Annotation at the leaves}: each leaf function is a
  separate cipher map.  The combiner chains them.  Intermediate cipher
  values are visible to the untrusted machine, leaking correlations
  between components (as in the product-type trade-off of
  \Cref{sec:products}).
\item \textbf{Annotation at intermediate nodes}: a spectrum between the
  two extremes.  Deeper annotations absorb more computation into cipher
  maps (more confidential, more space); shallower annotations expose
  more structure (less space, more leakage).
\end{itemize}

\begin{remark}[Shared variables and compositional leakage]
\label{rem:shared-vars}
When two cipher nodes $f$ and $g$ share an input variable $x$, the
untrusted machine observes two cipher values $\enc_f(x, k_f)$ and
$\enc_g(x, k_g)$ derived from the same latent $x$.
Even if both encodings are individually $\delta$-close to uniform, the
joint distribution of $(\enc_f(x), \enc_g(x))$ preserves the correlation
(as in \cite[Prop.~9.1]{towell2026cipher}).
The only mitigation is to merge $f$ and $g$ into a single cipher node
(eliminating the shared intermediate) or to inject noise queries that
dilute the correlation signal.
\end{remark}

\paragraph{Implementation via tracing.}
A practical realization uses Python function decorators and runtime
tracing.
Functions marked with a \texttt{@cipher\_node} decorator are
intercepted during a tracing pass over the input domain.
The tracing records which cipher nodes are called, with what arguments,
and what they return.
From this trace, the system:
\begin{enumerate}
\item builds a cipher map for each traced cipher node,
\item builds a combiner cipher map over the hash-output tuples of the
  components,
\item infers the input routing (which program arguments feed which
  cipher node) from the traced call patterns.
\end{enumerate}
Short-circuit evaluation (e.g., Python's \texttt{x if cond else y})
is handled by computing all component cipher maps for every input,
even when the original program skips some branches.
The combiner's domain includes hash tuples for both branches.
The resulting composed cipher program evaluates entirely on the
untrusted machine with no round trips to the trusted machine.


%% --------------------------------------------------------------------
\subsection{Cut-Point Structure Common to Both Realizations}
\label{sec:tm-vs-tree}
%% --------------------------------------------------------------------

Both realizations implement the same abstract move.

\begin{definition}[Cut point]
\label{def:cut-point}
Given a plaintext computation $P$ represented as a directed graph with
plaintext functions at nodes and data flow along edges, a \emph{cut
point} is a node whose subgraph of ancestors is replaced, during
construction, by a single cipher map.
Edges entering the cut point carry plaintext values on the trusted
machine and cipher values on the untrusted machine; edges leaving the
cut point carry cipher values throughout.
A \emph{cut-point set} is a collection of cut points such that every
input of the plaintext computation reaches at least one cut point, and
the cipher maps at the cut points together with a \emph{combiner}
cipher map for the nodes above them compute the full program on the
untrusted machine.
\end{definition}

\paragraph{Mapping the two realizations onto cut points.}
A cipher TM places a single cut point at the transition function
$\delta_T$, with a self-loop: the cut-point output $(c', d)$ feeds the
next step's tape cell, which feeds $\delta_T$ again.
Expression-tree decomposition places multiple cut points in a
finite-depth DAG, one at each \texttt{@cipher\_node}-annotated node,
with the combiner cipher map above them.
The number of cut points, their depth in the computation, and the
cipher spaces between them are the three design knobs.

\paragraph{What leaks above the cut points.}
Exactly what remains plaintext above the cut points is the leakage
profile.
For a cipher TM, this is the head-movement sequence (one clear direction
per step) and the halting time, nothing else.
For expression-tree decomposition, this is the collection of
intermediate cipher values flowing between cipher nodes, which leak
correlations whenever multiple cut points share latent inputs.

\paragraph{What the typed-chain bound says in each case.}
\Cref{prop:typed-orbit} bounds the adversary's orbit by $\sum_i N_i$
where the levels $i$ correspond to the depth of cut points along any
computation path.
For a cipher TM running for $T$ steps with self-loop at a single
(binary-arity, if state and symbol are distinct inputs) cut point, a
naive reading gives $N_i = 1$ for a single initial tape configuration
so the orbit is bounded by $1 + T$; with $m$ distinct initial tape
contents considered as a population, the bound is $m \cdot (T+1)$.
For an expression tree of depth $k$ with binary combinators, the bound
is $\sum_{i=0}^{k} m^{2^i}$ where $m$ is the number of distinct latent
inputs.
Typing the cipher spaces between cut points is what prevents the
self-loop in either realization from degenerating into an unbounded
orbit.

\begin{example}[Regex matching as cipher TM vs.\ expression tree]
\label{ex:regex-cut-points}
Consider matching a fixed regular expression $R$ of size $r$ against a
string $x \in \Sigma^\ell$.
As a cipher TM, $Q$ is the set of DFA states for $R$ (size $O(r)$),
$\Gamma = \Sigma \cup \{\sqcup\}$, and $\hat{\delta}_T$ is a single
cipher map of space $O(r \cdot |\Sigma|)$, evaluated $\ell$ times per
match.
Leakage: the head moves right exactly $\ell$ times, revealing $\ell$.
As an expression tree, the program is a fold over the string with $\ell$
cut points at the transition function; the combiner cipher map above
them has domain $\cipher{Q}^{\ell}$ so its construction cost is
exponential in $\ell$.
Cipher TM is preferable for long strings; expression-tree decomposition
is preferable when $\ell$ is small and string length must be hidden.
Choosing between the two is exactly a cut-point placement problem.
\end{example}


%% ====================================================================
\section{Cipher Boolean Algebra and Evaluation}
\label{sec:cipher-bool}
%% ====================================================================

The cipher Boolean type $\cipher{\mathrm{Bool}}$ specializes the
sum-type trade-off to the most fundamental algebraic structure and
provides the concrete instantiation we evaluate on a real corpus.


%% --------------------------------------------------------------------
\subsection{Cipher Boolean Type}
\label{sec:cipher-bool-type}
%% --------------------------------------------------------------------

\begin{definition}[Cipher Boolean type]
\label{def:cipher-bool}
A cipher Boolean type over $\B^n$ is a partition of the hash space into
three disjoint regions:
\begin{itemize}
\item $T \subset \B^n$: the True region ($|T| / 2^n = p_{\mathrm{true}}$),
\item $F \subset \B^n$: the False region ($|F| / 2^n = p_{\mathrm{false}}$),
\item $N = \B^n \setminus (T \cup F)$: the noise region.
\end{itemize}
The partition is derived from a secret via a pseudorandom permutation.
Without the secret, membership in $T$, $F$, or $N$ is indistinguishable
from random.
\end{definition}

\begin{remark}[Generalization beyond Bool]
\label{rem:general-cipher-type}
\Cref{def:cipher-bool} generalizes to any finite type $Y$.
A cipher type $\cipher{Y}$ partitions $\B^n$ into $|Y|$ value regions
plus a noise region, with $|R_y|/2^n \propto \Pr[Y = y]$
(Shannon-optimal allocation per the source coding
theorem~\cite{shannon1948mathematical}).
Operations $f : Y_1 \times \cdots \times Y_k \to Z$ become cipher maps
over the product of input regions.
The cipher Boolean is the special case $Y = \{\mathrm{True},
\mathrm{False}\}$; the same construction applies to enumerated types,
bounded integers, or any codomain where the output distribution is known.
\end{remark}


%% --------------------------------------------------------------------
\subsection{Noise Unreliability}
\label{sec:noise-unreliability}
%% --------------------------------------------------------------------

The noise region $N$ plays a structural role beyond mere error tolerance.

\begin{proposition}[Noise unreliability]
\label{prop:noise-unreliability}
Let $\mathrm{AND}$, $\mathrm{OR}$, $\mathrm{NOT}$ be cipher maps
implementing Boolean operations over the cipher Boolean type.
If any input is a noise value $n \in N$, the output is \emph{uncorrelated}
with the intended Boolean result: it lands in $T$, $F$, or $N$ according
to the allocation ratios $p_T$, $p_F$, $p_N$, independent of the other
input's value.
\end{proposition}

\begin{proof}
The AND, OR, and NOT cipher maps are constructed over the domain
$T \cup F$.
An input from $N$ is outside this domain.
Since the cipher map is total (backed by a PHF with random noise in
unused slots), a noise input maps to a random slot whose contents decode
to a value in $T$ with probability $p_T$, in $F$ with probability $p_F$,
and in $N$ with probability $p_N$.
For the standard allocation ($p_T = 0.05$, $p_F = 0.90$, $p_N = 0.05$),
a noise input produces False 90\% of the time, True 5\%, and noise 5\%.
Crucially, this output is independent of the other (valid) input:
the cipher map cannot ``correct'' for a noise input because it does not
know which inputs are noise.
\end{proof}

\begin{remark}[Noise prevents output certainty]
The noise region ensures that an adversary observing a sequence of
Boolean cipher map evaluations cannot achieve certainty about any
individual result.
Consider an adversary that repeatedly ANDs a cipher value $c$ with
random cipher values $r_1, r_2, \ldots$.
If $c \in T$ (True), then $\mathrm{AND}(c, r_i)$ produces True when
$r_i \in T$ and False when $r_i \in F$.
But when $r_i \in N$ (5\% probability), the output is noise, which the
adversary cannot distinguish from a legitimate False.
After $k$ ANDs, the adversary has seen some True and some False/noise
results, but cannot determine with certainty whether any given False
result is real or noise-induced.

Dually, an adversary that repeatedly ORs $c$ with random values sees
True for $c \in T$ (always True), but again noise inputs produce
unpredictable outputs.
For $c \in F$, the OR produces True only when $r_i \in T$ (5\%),
False when $r_i \in F$ (90\%), and noise when $r_i \in N$ (5\%).
The noise fraction creates a floor of uncertainty in any statistical
test the adversary performs.
\end{remark}

\paragraph{Practical allocation.}
The cipher Boolean allocation $|T|/2^n = 0.05$, $|F|/2^n = 0.90$,
$|N|/2^n = 0.05$ is motivated by the set membership use case.
A cipher set $\hat{1}_A : \cipher{X} \to \cipher{\mathrm{Bool}}$
maps domain elements to cipher Booleans.
Members are forced into $T$ by the construction.
Non-members land randomly: 90\% decode as False (correct rejection),
5\% as True (false positive), 5\% as noise.
The false positive rate equals $p_{\mathrm{true}}$ and is a system
parameter, not dependent on the set size or universe.

\paragraph{Composition with set operations.}
Boolean search queries like ``$w_1 \land w_2 \land \lnot w_3$'' are
evaluated entirely on the untrusted machine by chaining cipher set
evaluations through cipher AND/OR/NOT maps.
Each Boolean operation is itself a cipher map (constructed via PHF
over the $|T \cup F|^2$ input pairs for AND/OR, or $|T \cup F|$
for NOT).
The untrusted machine sees opaque hash values flowing through opaque
total functions at every stage.


%% --------------------------------------------------------------------
\subsection{Experimental Validation}
\label{sec:evaluation}
%% --------------------------------------------------------------------

A reference implementation (\texttt{cipher-maps}, Python with PHF
backend via \texttt{phobic}) validates the construction on the 20
Newsgroups corpus (18{,}266 documents, 58{,}903 unique words).
With $n = 8$ bits, $|T| = 13$, $|F| = 230$, the AND and OR cipher
maps each have $(13 + 230)^2 = 59{,}049$ entries.

\paragraph{Boolean search at scale.}
\Cref{tab:boolean-search} shows precision and recall for
representative Boolean queries over 5{,}000 documents, with all
Boolean operations evaluated on the untrusted machine.
Construction rate: 843 documents per second.

\begin{table}[ht]
\centering
\caption{Boolean encrypted search (5{,}000 documents, $p_T = 0.05$).}
\label{tab:boolean-search}
\begin{tabular}{@{}lcccc@{}}
\toprule
Query & Precision & Recall & FP & Time \\
\midrule
Single term & 0.39 & 1.00 & 248 & 0.10s \\
2-term AND & 0.26 & 1.00 & 42 & 0.30s \\
3-term AND & 0.33 & 1.00 & 12 & 0.60s \\
OR & 0.37 & 0.97 & 428 & 0.30s \\
OR AND NOT & 0.33 & 0.88 & 389 & 0.55s \\
\bottomrule
\end{tabular}
\end{table}

AND queries maintain perfect recall while false positives decrease
(248 for single term, 12 for 3-term AND).
OR and NOT queries lose recall (0.97 and 0.88) due to noise
propagation: when a cipher set returns a noise value $n \in N$,
the subsequent OR or NOT operation produces noise or an unpredictable
result, effectively dropping the document from the result set.

\paragraph{FPR compounding vs.\ Bernoulli model.}
The Bernoulli composition model~\cite{towell2026bernoulli} predicts
that $k$ independent AND operations yield
$\mathrm{FPR}_k = p_T^k$.
Empirically, OR chains match theory closely (ratio $\approx 0.9$
across chain lengths 1--5), but AND chains diverge at depth $> 2$:
the empirical FPR exceeds $p_T^k$ by an order of magnitude at
$k = 3$.
The reason: each Bernoulli test is an independent draw, but cipher
Boolean AND is a deterministic cipher map.
After one AND, the result is a specific cipher Boolean value, and
feeding it into the next AND produces a deterministic output.
The independence assumption breaks at the cipher map level even
though it holds within a single cipher set.

\paragraph{Encoding granularity.}
We built a 7-function loan approval pipeline at three granularity
levels:
\begin{center}
\begin{tabular}{@{}lccc@{}}
\toprule
Level & Build time & Space & Intermediates \\
\midrule
Root (1 cipher map) & 0.009s & 694\,B & 0 \\
Intermediate (3 groups) & 6.8s & --- & 3 \\
Leaf (7 nodes) & 5.9s & --- & 7 \\
\bottomrule
\end{tabular}
\end{center}
All three produce zero errors on the full domain (150 inputs).
The root annotation is 650$\times$ faster (PHF backend vs.\ seed
search) and exposes zero intermediate cipher values.
The trade-off: the root annotation requires the entire input domain
at construction time, while the leaf annotation builds smaller cipher
maps that compose dynamically.


%% ====================================================================
\section{Discussion}
\label{sec:discussion}
%% ====================================================================

\paragraph{Relationship to cipher maps.}
This paper extends the cipher map framework of~\cite{towell2026cipher} in
two directions: inward (what happens when cipher values carry algebraic
structure) and outward (what happens when the untrusted machine actively
explores the cipher space).
The cipher map paper focuses on individual maps and their four properties;
this paper studies how those properties interact with type constructors and
how confidentiality degrades under active use.

\paragraph{Relationship to Bernoulli data types.}
The accuracy-side analysis of algebraic types over approximate values is
developed in~\cite{towell2026bernoulli}, which studies the Kronecker
factorization of joint confusion matrices for product types and the error
propagation through sum-type constructors.
That paper asks: how does $\eta$ change when types are composed?
This paper asks the complementary question: how does confidentiality
change when types are composed?
Together, they characterize the full trade-off space (accuracy $\times$
confidentiality) for algebraic types over trapdoor computing.

\paragraph{The encoding granularity principle.}
The product and sum trade-offs in \Cref{sec:products,sec:sums} are
instances of a single principle developed
in~\cite[Sec.~9]{towell2026cipher}: coarser encoding granularity
(encoding more values as a single unit) yields better confidentiality but
fewer operations available to the untrusted machine.
The entanglement parameter $p$~\cite[Sec.~9.3]{towell2026cipher} controls
the spectrum: $p = 1$ (component-wise) gives maximum functionality and
minimum confidentiality; $p = k$ (whole-state) gives maximum
confidentiality and minimum functionality.

\paragraph{Open questions.}

\begin{enumerate}
\item \emph{Recursive types.}
  Lists, trees, and other recursive types are built from sums and products.
  The sum-type impossibility (\Cref{thm:sum-impossibility}) implies that
  recursive traversal (which requires pattern matching at each node)
  either leaks the structure (which constructors are used at each level)
  or requires a trusted-machine round trip at each node.
  Quantifying the total leakage for a recursive traversal of depth $d$
  is open.

\item \emph{Tight orbit bounds for specific constructions.}
  \Cref{thm:confidentiality-bound} gives a general upper bound on
  confidentiality in terms of orbit size.
  For specific cipher map constructions (e.g., the trapdoor boolean
  algebra of~\cite{towell2026cipher}), tighter bounds may be possible
  by exploiting the structure of the construction.

\item \emph{Cipher TM with oblivious head.}
  Can the head direction be ciphered at acceptable cost?
  The naive approach (evaluate all three possible next states) triples
  the work per step.
  Whether a sub-multiplicative overhead exists, achieving partial
  head-movement hiding without full ORAM cost, is open.

\item \emph{Partial sum-type hiding.}
  \Cref{thm:sum-impossibility} shows that full tag hiding and full
  untrusted pattern matching are mutually exclusive.
  Is there a meaningful intermediate notion---hiding the tag from
  some adversary classes while permitting restricted pattern matching?
  For instance, one could imagine a construction where the tag is hidden
  against passive adversaries (who only observe cipher values) but
  available to active adversaries (who evaluate cipher maps).

\item \emph{Cut-point placement as an optimization.}
  \Cref{sec:tm-vs-tree} describes cut points as a design knob spanning
  cipher-TM-style iteration and leaf-level expression-tree annotation.
  Formulating cut-point selection as a constrained optimization, subject
  to the orbit bound of \Cref{prop:typed-orbit} and construction-time
  space budgets, would turn the choice into a solvable problem rather
  than a designer's judgement call.
\end{enumerate}


%% ====================================================================
\section{Conclusion}
%% ====================================================================

We have shown that algebraic type constructors create unavoidable
confidentiality costs in trapdoor computing.
The sum-type impossibility theorem establishes that tag hiding and
untrusted pattern matching are mutually exclusive for any encoding.
The orbit closure framework provides a quantitative bound on
confidentiality degradation as the untrusted machine applies
operations, and a typed-composition discipline converts that bound
into a design-time budget.
These results apply to any cipher map construction, not just specific
implementations.

Two realizations concretize these trade-offs.
Expression-tree decomposition with cipher-node annotations lets the
programmer choose the granularity at which intermediate cipher values
are exposed; a cipher Turing machine gives space $O(|Q| \cdot |\Gamma|)$
at the cost of leaking head-movement patterns.
Both are instances of a single cut-point pattern that the typed-chain
bound constrains identically.

On the practical side, cipher Boolean types with AND/OR/NOT as cipher
maps enable Boolean search queries evaluated entirely on the untrusted
machine.
Experiments on the 20 Newsgroups corpus (5{,}000 documents) show
perfect recall for AND queries, with false positive rates matching
theoretical predictions for single-term queries and compounding
predictably through Boolean chains.
The encoding granularity spectrum is measurable: root annotation
(one cipher map for the entire program) is 650$\times$ faster and
exposes zero intermediates compared to leaf annotation
(one cipher map per function).
The gap between the Bernoulli independence model and the cipher map
implementation for AND chain composition points to future work on
tighter error models for composed cipher maps.

%% ====================================================================
\section*{Acknowledgments}
%% ====================================================================

The algebraic cipher types concept originates in a 2019--2022 C++ notebook
by the author, which explored cipher Booleans, cipher unions, and cipher
Turing machines through concrete implementations.
The present paper extracts and formalizes the theoretical content from that
notebook, building on the cipher map framework developed
in~\cite{towell2026cipher}.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
