\documentclass[11pt]{article}

% Essential packages
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage[numbers,square]{natbib}
\usepackage{booktabs}
\usepackage{hyperref}
\hypersetup{hypertexnames=false}
\usepackage{cleveref}

% Theorem environments (matching cipher-maps paper)
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{remark}
\newtheorem{remark}{Remark}[section]
\newtheorem{example}{Example}[section]

% Notation (matching cipher-maps paper)
\newcommand{\fhat}{\hat{f}}
\newcommand{\ghat}{\hat{g}}
\newcommand{\enc}{\mathrm{enc}}
\newcommand{\dec}{\mathrm{dec}}
\newcommand{\im}{\mathrm{Im}}
\newcommand{\TV}{d_{\mathrm{TV}}}
\newcommand{\B}{\{0,1\}}
\DeclareMathOperator{\Uniform}{Uniform}
\newcommand{\cipher}[1]{\mathsf{C}(#1)}
\DeclareMathOperator{\orbitF}{orbit}

\title{The Entropy Ratio: Quantitative Confidentiality\\for Trapdoor Computing}
\author{Alexander Towell\\\texttt{lex@metafunctor.com}}
\date{April 2026}

\begin{document}

\maketitle

\begin{abstract}
A cipher map is a total function on bit strings implementing a trapdoor
approximation of a latent function: the untrusted machine evaluates it
blindly; only the trusted machine, holding the secret, can encode inputs
and decode outputs.  We develop a quantitative confidentiality theory
for cipher map systems.  The central measure is the \emph{entropy
ratio} $e = H/H^*$, a normalized form of Shannon leakage from the
Quantitative Information Flow literature.  Our contribution is a bridge
from the cipher map framework to this measure: the
representation-uniformity parameter $\delta$ controls the ratio via the
Fannes--Audenaert continuity inequality, giving
$e \geq 1 - \delta - h_2(\delta)/n$, which reduces confidentiality
engineering to the concrete problem of minimizing $\delta$.  We analyze
two constructions that reduce $\delta$ with explicit costs (noise
injection and multiple representations), and inherit a third, encoding
granularity, from the cipher maps framework.  We prove a compositional
leakage theorem: even when marginals are $\delta$-uniform, the joint
distribution over shared cipher values is recoverable at standard
parametric rate, showing that reducing $\delta$ is necessary but not
sufficient.  Experimental validation on encrypted Boolean search over
the 20 Newsgroups corpus confirms the theoretical predictions for FPR
compounding and the encoding granularity spectrum.
\end{abstract}


%% ====================================================================
\section{Introduction}
\label{sec:intro}
%% ====================================================================

Given a cipher map system---a trusted machine encoding queries, an
untrusted machine evaluating total functions on bit strings, and results
decoded only by the trusted machine---how much does the untrusted machine
learn?

The cipher map framework~\cite{towell2026cipher} provides four properties
(totality, representation uniformity, correctness, composability)
parameterized by $(\eta, \varepsilon, \delta, \mu)$ that characterize
what the untrusted machine can and cannot do.  Totality ensures every
input produces output; representation uniformity bounds frequency
analysis; correctness governs error rates; composability enables chained
evaluation.  But these properties are \emph{qualitative} guarantees.  A
system designer needs \emph{quantitative} answers: given a specific
deployment with a known query distribution, known vocabulary size, and
known resource budget, how confidential is the system, and how can it be
improved?

This paper develops the quantitative theory.  The central idea is the
\emph{entropy ratio} $e = H/H^*$: the ratio of observed entropy in the
cipher map output stream to the maximum entropy achievable under system
constraints.  The measure itself is not new; it is a normalized form of
Shannon leakage studied in the Quantitative Information Flow literature
(\S\ref{sec:related}).  The contribution here is the connection to the
cipher map framework: we prove that the representation-uniformity
parameter $\delta$ directly lower-bounds the entropy ratio via the
Fannes--Audenaert continuity
inequality~\cite{fannes1973continuity,audenaert2007sharp},
\[
e \;\geq\; 1 - \delta - h_2(\delta)/n,
\]
where $h_2$ is the binary entropy.  This reduces confidentiality
engineering to the concrete problem of minimizing $\delta$.

We analyze two constructions that reduce $\delta$ with explicit costs:

\begin{enumerate}
\item \textbf{Noise injection.}  The trusted machine mixes $R$ filler
  queries per $N$ real queries.  By totality, the untrusted machine
  cannot distinguish real from filler.  Cost: bandwidth
  ($1 + R/N$).  Effect: Fisher information for any correlation
  estimator is reduced by $\rho^2 = (N/(N+R))^2$.

\item \textbf{Multiple representations.}  Each element receives
  $K(x) \geq 1$ encodings, with $K(x) \propto D(x)$ to flatten the
  cipher value distribution (classical homophonic substitution:
  frequent elements get more code symbols so that per-cipher frequency
  equalizes).  Cost: space ($\sum_x K(x)$ slots).  Effect:
  $\delta \to 0$ as the normalizing constant grows.
\end{enumerate}

A third construction, \textbf{encoding granularity} (controlled by an
entanglement parameter $p$ that interpolates between component-wise
encoding with correlation leakage and joint encoding with space
$O(|Y|^k)$), is inherited from the cipher maps framework
\cite[Sec.~9]{towell2026cipher}.  We use it but do not re-prove it
here.

The paper's compositional contribution is distinct from either
construction.  We prove (Theorem~\ref{thm:comp-leakage}) that
$\delta$-uniform marginals do not prevent joint recovery when the
untrusted machine observes multiple evaluations on a shared cipher
value: the joint distribution is recoverable at standard parametric
rate $O(|Y_1|\cdot |Y_2|/\xi^2)$.  This is the reason practitioners
cannot stop at multiplicity.  We also include, for completeness, FPR
compounding through Boolean chains, which is inherited
from~\cite[Sec.~8]{towell2026cipher}.

Finally, we provide practical measurement tools.  The entropy of a cipher
map output stream can be estimated by compressing it: more compressible
means more predictable means less confidential.  This avoids the need for
explicit probabilistic models of the adversary.

\paragraph{Contributions.}
\begin{enumerate}
\item \textbf{Fannes bridge.}  The representation-uniformity parameter
  $\delta$ directly lower-bounds the entropy ratio,
  $e \geq 1 - \delta - h_2(\delta)/n$, via the Fannes--Audenaert
  continuity inequality (Theorem~\ref{thm:entropy-decomposition}).
  This makes $\delta$ the operational handle for confidentiality
  engineering in cipher map systems (\S\ref{sec:measure}).

\item \textbf{Two constructions that reduce $\delta$.}  Noise injection
  (bandwidth cost $1 + R/N$, Fisher-information dilution $\rho^2$;
  Theorem~\ref{thm:noise-dilution}) and multiple representations
  (space cost $\sum_x K(x)$ with $K(x)\propto D(x)$, the classical
  homophonic prescription; Theorem~\ref{thm:multiplicity}).  A third
  construction, encoding granularity, is inherited from the cipher
  maps framework and used without re-proof (\S\ref{sec:levers}).

\item \textbf{Compositional leakage bound.}
  Theorem~\ref{thm:comp-leakage} shows that $\delta$-uniform marginals
  do not prevent joint recovery when the untrusted machine observes
  multiple evaluations on a shared cipher value: the joint distribution
  is recoverable from $O(|Y_1|\cdot |Y_2|/\xi^2)$ observations.  This
  is the paper's main compositional contribution and is the reason
  reducing $\delta$ is necessary but not sufficient.  FPR compounding
  through Boolean chains is included for context; it is inherited
  from~\cite{towell2026cipher} (\S\ref{sec:composition}).

\item \textbf{Practical measurement and validation.}  Compression-based
  entropy estimation avoiding explicit adversary models, and a case
  study demonstrating confidentiality improvement from 72\% to 98\%
  with approximately 1.04$\times$ space and 1.5$\times$ bandwidth
  overhead (\S\ref{sec:measurement},~\S\ref{sec:experiments}).
\end{enumerate}

\paragraph{What this paper does not do.}
We do not use ORAM, FHE, or simulation-based security definitions.
Privacy in the cipher map model comes from the one-way hash, the totality
of $\fhat$, and representation uniformity---not from access-pattern
indistinguishability or computational hardness assumptions beyond the
hash function.  The framework is information-theoretic throughout.


%% ====================================================================
\section{Related Work}
\label{sec:related}
%% ====================================================================

\paragraph{Quantitative information flow.}
The entropy ratio $e = H/H^*$ is a normalized form of Shannon leakage,
studied extensively in the QIF
literature~\cite{smith2009foundations,alvim2020science}.
Smith~\cite{smith2009foundations} proposes min-entropy leakage as a
worst-case measure; Alvim et al.~\cite{alvim2020science} provide a
comprehensive treatment of information-theoretic leakage measures.
The measure itself is not our contribution, and we do not claim
novelty there.  What is new here is (i) the \emph{Fannes bridge}
connecting the representation-uniformity parameter $\delta$ of cipher
maps to the entropy ratio ($e \geq 1 - \delta - h_2(\delta)/n$),
making $\delta$ the operational handle QIF does not have, and
(ii) the compositional leakage bound showing that $\delta$-uniform
marginals are insufficient when shared variables recur across
evaluations (Theorem~\ref{thm:comp-leakage}).  The QIF literature
treats leakage of a function; the cipher map setting adds a
four-parameter construction $(\eta, \varepsilon, \delta, \mu)$ and a
composition discipline under which that leakage must be reasoned
about.  We discuss the Shannon-vs-min-entropy choice in
\S\ref{sec:discussion}.

\paragraph{SSE leakage mitigation.}
Noise injection for encrypted search has been studied in the SSE
literature.
Bost and Fouque~\cite{bost2017thwarting} propose dummy queries with
game-based security analysis; Demertzis et al.~\cite{demertzis2020seal}
develop tunable leakage-functionality trade-offs.
Our analysis differs in using information-theoretic measures (entropy
ratio) rather than simulation-based definitions, and applies to general
cipher maps, not only encrypted search.

\paragraph{Leakage from property-preserving encryption.}
Naveed et al.~\cite{naveed2015inference} and Islam et
al.~\cite{islam2012access} demonstrate inference attacks exploiting
preserved algebraic structure.
The cipher map model avoids property preservation entirely: the untrusted
machine evaluates a total function on opaque bit strings and cannot
determine the output type from the cipher value.

%% ====================================================================
\section{Preliminaries}
\label{sec:prelim}
%% ====================================================================

We recall the cipher map abstraction from~\cite{towell2026cipher}.  All
notation follows that paper; we cite rather than re-derive.

\begin{definition}[Cipher map~{\cite[Def.~3.1]{towell2026cipher}}]
\label{def:cipher-map}
A \emph{cipher map} for a latent function $f : X \to Y$ is a tuple
$(\fhat, \enc, \dec, s)$ where $\fhat : \B^n \to \B^n$ is a total
function on $n$-bit strings, $\enc : X \times \{0,\ldots,K(x){-}1\}
\to \B^n$ is an encoding function with $K(x) \geq 1$ representations
per element, $\dec : \B^n \to Y \cup \{\bot\}$ is a decoding function,
and $s$ is a secret (the trapdoor).  The untrusted machine holds $\fhat$.
The trusted machine holds $\enc$, $\dec$, and $s$.
\end{definition}

The four properties, parameterized by
$(\eta, \varepsilon, \mu, \delta)$~\cite[Sec.~4]{towell2026cipher}:

\begin{description}
\item[Totality.]  $\fhat$ is defined on all $2^n$ inputs.  Out-of-domain
  outputs are indistinguishable from uniform under the random oracle
  model.

\item[Representation uniformity ($\delta$-bounded).]  The marginal
  distribution of cipher values is $\delta$-close to uniform over $\B^n$
  in total variation distance:
  $\TV(P, U) = \frac{1}{2}\sum_c |P(c) - U(c)| \leq \delta$.
  Informally, $\delta$ bounds the advantage of any test that tries to
  distinguish real cipher values from random ones.

\item[Correctness ($\eta$-bounded).]  The fraction of domain elements
  that decode incorrectly is at most $\eta$:
  \[
  \Pr[\dec(\fhat(\enc(x,k))) \neq f(x)] \leq \eta.
  \]
  Nonzero $\eta$ provides plausible deniability (an observed result might
  be wrong) and reduces construction cost.

\item[Noise-decode probability ($\varepsilon$).]  A random $n$-bit
  string decodes to a valid output with probability $\varepsilon$.
  Lower $\varepsilon$ means out-of-domain queries are more likely to be
  rejected (decoded as $\bot$), at the cost of more bits per element.

\item[Value cost ($\mu = H(Y)$).]  The average number of bits needed to
  encode one output value, equal to the Shannon entropy of the output
  distribution.

\item[Composability.]  The composition of two cipher maps is again a
  cipher map: $\ghat \circ \fhat$ has correctness
  $\eta_{g \circ f} \leq 1 - (1 - \eta_f)(1 - \eta_g)$.
  Errors accumulate predictably, enabling pipeline analysis.
\end{description}

The space per element decomposes as
$\text{bits/element} = -\log_2 \varepsilon + H(Y)$, where
$-\log_2 \varepsilon$ bits distinguish signal from noise and
$H(Y) = \mu$ bits encode the function
value~\cite[Thm.~6.1]{towell2026cipher}.

\paragraph{Trusted/untrusted model.}
The trusted machine $T$ encodes inputs, decodes results, and injects
filler queries.  The untrusted machine $U$ evaluates $\fhat$ on any bit
string and returns results.  $U$ cannot decode, distinguish real from
filler queries, determine the domain $X$, or enumerate which inputs are
``in-domain''~\cite[Sec.~5]{towell2026cipher}.

\paragraph{Acceptance predicates.}
The batch construction of a cipher map is unified under an
\emph{acceptance predicate}: a family of disjoint sets
$\{A(y) \subseteq \B^n : y \in Y\}$ that partitions hash space among
output values~\cite[Sec.~6.2]{towell2026cipher}.  Shannon-optimal
allocation sets $|A(y)| \propto \Pr[f(X) = y]$, simultaneously
minimizing space and maximizing output indistinguishability.

\paragraph{Information-theoretic notation.}
We use standard information-theoretic quantities throughout.
\emph{Shannon entropy} $H(X) = -\sum_x p(x) \log_2 p(x)$ measures the
average surprise (in bits) of a random variable; higher entropy means
less predictable.
\emph{Conditional entropy} $H(X \mid Y)$ is the remaining uncertainty in
$X$ after observing $Y$.
\emph{Mutual information} $I(X; Y) = H(X) - H(X \mid Y)$ quantifies how
much observing $Y$ reveals about $X$; zero means independence.
\emph{KL divergence} $D_{\mathrm{KL}}(P \| Q) = \sum_x P(x) \log_2
(P(x)/Q(x))$ measures how far $P$ is from $Q$; it is zero iff $P = Q$
and connects to the entropy ratio via $H(P) = H^* - D_{\mathrm{KL}}(P
\| U)$ when $U$ is the maximum-entropy (uniform) distribution.
We use $D$ for the distribution of queries over the domain $X$, and $Q$
for the induced distribution over cipher values.
All logarithms are base~2.


%% ====================================================================
\section{The Confidentiality Measure}
\label{sec:measure}
%% ====================================================================

\subsection{Observed and Maximum Entropy}

Consider a cipher map system in which the trusted machine submits a
stream of cipher values to the untrusted machine.  The untrusted machine
observes a sequence $c_1, c_2, \ldots, c_N$ of $n$-bit strings, each
either a real encoding $\enc(x_i, k_i)$ or a filler string drawn
uniformly from $\B^n$.

\begin{definition}[Observed entropy]
\label{def:observed-entropy}
The \emph{observed entropy} $H$ of a cipher map system is the Shannon
entropy of the joint distribution over the observable sequence:
\begin{equation}
H = H(C_1, C_2, \ldots, C_N),
\end{equation}
where each $C_i$ is the random variable corresponding to the $i$-th
cipher value observed by the untrusted machine.
\end{definition}

\begin{definition}[Maximum entropy under constraints]
\label{def:max-entropy}
The \emph{maximum entropy} $H^*$ is the maximum of $H$ over all
distributions on the observable sequence that are consistent with the
system constraints: domain size $|X|$, codomain size $|Y|$, and the
cipher map parameters $(\eta, \varepsilon, \delta)$.
\end{definition}

By the maximum entropy principle~\cite{jaynes1957information}, $H^*$ is
achieved when the observable distributions are as uniform as possible
subject to the constraints.  For a single cipher value drawn from a
vocabulary of size $m$, the maximum entropy is $\log_2 m$ (uniform
distribution).  For a sequence of $N$ independent cipher values, the
maximum entropy is $N \log_2 m$.

\subsection{The Entropy Ratio}

\begin{definition}[Entropy ratio]
\label{def:entropy-ratio}
The \emph{entropy ratio} of a cipher map system is
\begin{equation}
\label{eq:entropy-ratio}
e = \frac{H}{H^*} \in [0, 1].
\end{equation}
\end{definition}

The entropy ratio has a direct operational interpretation:

\begin{itemize}
\item $e = 0$: the output sequence is fully predictable.  The untrusted
  machine can predict every cipher value with certainty.  No
  confidentiality.

\item $e = 1$: the output sequence is indistinguishable from the maximum
  entropy distribution under system constraints.  The untrusted machine's
  best prediction is no better than guessing from the maximum entropy
  distribution.  Maximum confidentiality.
\end{itemize}

\subsection{Relationship to Cipher Map Parameters}

The entropy ratio connects to the four cipher map parameters through the
following result.

\begin{theorem}[Entropy ratio decomposition]
\label{thm:entropy-decomposition}
Let $(\fhat, \enc, \dec, s)$ be a cipher map for $f : X \to Y$ with
parameters $(\eta, \varepsilon, \delta, \mu)$.  Let $D$ be the query
distribution on $X$, and let $Q$ be the induced distribution on cipher
values.  Then the per-query observed entropy satisfies:
\begin{equation}
\label{eq:entropy-decomp}
H(Q) = H^* - D_{\mathrm{KL}}(Q \| U)
      = H^* \!\left(1 - \frac{D_{\mathrm{KL}}(Q \| U)}{H^*}\right),
\end{equation}
where $U$ is the uniform distribution on $\B^n$ and the entropy ratio is
$e = H(Q)/H^* = 1 - D_{\mathrm{KL}}(Q \| U)/H^*$.  In particular:
\begin{enumerate}
\item If $\delta = 0$ (perfect representation uniformity), then $Q = U$
  and $H(Q) = H^* = n$ bits, giving $e = 1$.
\item If $K(x) = 1$ for all $x$ (simple substitution, no multiple
  representations), then $Q$ inherits the non-uniformity of $D$, and
  $e \leq H(D) / \log_2 |X|$.
\item For $\delta \leq 1/2$, the entropy ratio is bounded by
  \[
  e \geq 1 - \delta - h_2(\delta)/n,
  \]
  where $h_2(\delta) = -\delta\log_2\delta - (1-\delta)\log_2(1-\delta)$
  is the binary entropy.  The bound follows from the Fannes--Audenaert
  continuity inequality~\cite{fannes1973continuity,audenaert2007sharp}.
\end{enumerate}
\end{theorem}

\begin{proof}
For part (1): when $\delta = 0$, $Q$ is uniform over $\B^n$ by
definition of representation uniformity.  The entropy of the uniform
distribution on $2^n$ elements is $n$ bits, which equals $H^*$.

For part (2): with $K(x) = 1$, the encoding $\enc$ is injective, so $Q$
is a relabeling of $D$.  Entropy is invariant under bijection, giving
$H(Q) = H(D)$.  The maximum over distributions on $|X|$ elements is
$\log_2 |X|$, so $e = H(D) / \log_2 |X|$.

For part (3): the Fannes--Audenaert inequality bounds entropy
continuity: for any distributions $P, Q$ on a support of size $d$ with
$\TV(P, Q) \leq t \leq 1/2$,
\[
|H(P) - H(Q)| \;\leq\; t \log_2(d - 1) + h_2(t).
\]
Applying this to $Q$ and $U$ on $\B^n$ with $d = 2^n$ and
$\TV(Q, U) \leq \delta$,
\[
|H(Q) - n| \;\leq\; \delta \log_2(2^n - 1) + h_2(\delta)
                \;\leq\; \delta n + h_2(\delta),
\]
so $H(Q) \geq n(1 - \delta) - h_2(\delta)$ and
$e = H(Q)/n \geq 1 - \delta - h_2(\delta)/n$.
Note the bound is linear in $\delta$ (not quadratic):
Pinsker gives $D_{\mathrm{KL}} \geq 2\,\TV^2$, an upper bound on TV
given KL, which does not convert to a KL bound given TV.  The correct
direction is Fannes--Audenaert.
\end{proof}

\begin{remark}[Numerical scale]
For $\delta = 0.05$ and $n = 64$, the bound gives $e \geq 0.945$.  For
$\delta = 0.01$, $e \geq 0.989$.  The linear scaling in $\delta$ means
tightening $\delta$ by an order of magnitude tightens the leakage bound
by roughly the same order.
\end{remark}

\begin{remark}[The role of $\varepsilon$]
The noise decode probability $\varepsilon$ does not appear directly in
the per-query entropy, but it determines the space cost: achieving small
$\varepsilon$ requires $-\log_2 \varepsilon$ bits per element.  The
indirect effect is that larger $\varepsilon$ means a denser cipher space
(more random bit strings decode to valid outputs), which helps uniformity
but increases false positives.
\end{remark}

\begin{remark}[The role of $\eta$]
The correctness parameter $\eta$ provides a form of plausible
deniability.  With $\eta > 0$, the untrusted machine cannot be certain
that a decoded output reflects the true function value.  This adds
$H_b(\eta) = -\eta \log_2 \eta - (1-\eta) \log_2(1-\eta)$ bits of
uncertainty per query (the binary entropy of the error event).
\end{remark}


%% ====================================================================
\section{Three Levers for Improving Confidentiality}
\label{sec:levers}
%% ====================================================================

\subsection{Noise Injection}
\label{subsec:noise}

The trusted machine can mix $R$ filler queries with $N$ real queries.  By
totality, the untrusted machine cannot distinguish real from filler: both
are $n$-bit strings, and $\fhat$ produces $n$-bit output on both.

\begin{theorem}[Noise dilution]
\label{thm:noise-dilution}
Assume $K(x) = 1$ (single encoding per element), so the cipher value
distribution of real queries equals $D$.
Let the trusted machine submit $N$ real queries drawn from $D$ and
$R$ filler queries drawn uniformly from $\B^n$.  The mixed
sequence has $N + R$ elements, each of which is real with probability
$\rho = N / (N + R)$ and filler with probability $1 - \rho$.  Then:
\begin{enumerate}
\item The per-element entropy of the mixed distribution is
\begin{equation}
\label{eq:noise-entropy}
H_{\mathrm{mix}} = H_b(\rho) + \rho \cdot H(D) + (1 - \rho) \cdot n,
\end{equation}
where $H_b(\rho)$ is the binary entropy of $\rho$.

\item As $R/N \to \infty$, $H_{\mathrm{mix}} \to n = H^*$ and $e \to 1$.

\item The adversary's estimate of any correlation between query elements
  has variance $O(1/(N+R))$.  Mixing with $R \gg N$ filler dilutes the
  real signal proportionally: the adversary needs
  $\Omega((1 + R/N)^2)$ observations to achieve the same estimation
  accuracy as observing $N$ pure real queries.
\end{enumerate}
\end{theorem}

\begin{proof}
For part (1): each observed element is a mixture.  With probability
$\rho$ it is drawn from $D$ (entropy $H(D)$); with probability
$1 - \rho$ it is uniform on $\B^n$ (entropy $n$).  By the chain rule,
the per-element entropy is the entropy of the mixing indicator plus the
conditional entropy given the indicator.

For part (2): as $\rho \to 0$, the distribution converges to uniform on
$\B^n$.

For part (3): the adversary observes a mixture.  For any statistic
$T$ computed from the observed sequence, the contribution of real
queries is diluted by the noise fraction.  Specifically, let
$\hat{\theta}$ be the adversary's estimator of a parameter $\theta$ of
$D$.  Each observation contributes information about $\theta$ only when
it is a real query (probability $\rho$).  The effective sample size for
estimating $\theta$ is $\rho \cdot (N + R) = N$, but the adversary does
not know which observations are real.  The best the adversary can do is
treat each observation as drawn from the mixture.
When $\rho \ll 1$ (filler dominates), the mixture is approximately
uniform and the Fisher information for estimating any parameter of $D$
is reduced by a factor of approximately $\rho^2$ relative to observing
$D$ directly.
\end{proof}

\paragraph{Cost.}
Noise injection costs bandwidth: the total query volume increases by a
factor of $(N + R) / N = 1 + R/N$.  It does not cost space (no
additional cipher map storage).  A noise ratio of $R/N = 1$ (equal
filler and real queries) doubles bandwidth; $R/N = 9$ (90\% filler)
increases bandwidth tenfold but brings $e$ close to 1 for most practical
distributions.

\subsection{Multiple Representations ($K > 1$)}
\label{subsec:multiple-reps}

Each element $x$ can be given $K(x) \geq 1$ distinct encodings.  The
trusted machine, when encoding $x$, chooses $k$ uniformly from
$\{0, \ldots, K(x) - 1\}$, so the probability of observing a particular
cipher value $v = \enc(x, k)$ is $D(x)/K(x)$ (assuming distinct
encodings, which holds with high probability under the random oracle
model).  Flattening the cipher value distribution therefore requires
$D(x)/K(x)$ to be constant in $x$: more frequent elements need
\emph{more} representations, not fewer.  This is the classical
homophonic substitution prescription~\cite{simmons1979symmetric}:
$K(x) \propto D(x)$.

\begin{theorem}[Representation uniformity via multiplicity]
\label{thm:multiplicity}
Let $D$ be a distribution on $X$ with $D(x) > 0$ for all $x$, and let
$c \geq 1/\min_x D(x)$ be a normalizing constant.  Set
$K(x) = \lceil c \cdot D(x) \rceil$.  Then the cipher value
distribution $Q$ induced on the image of $\enc$ satisfies
\begin{equation}
\TV(Q, U_{\mathrm{im}}) \;\leq\; \frac{|X|}{2\,\sum_x K(x)}
\;\leq\; \frac{|X|}{2c},
\end{equation}
where $U_{\mathrm{im}}$ is the uniform distribution on
$\mathrm{im}(\enc)$.  In particular, $\TV(Q, U_{\mathrm{im}}) \to 0$
as $c \to \infty$.  Under the random oracle model,
$\mathrm{im}(\enc)$ is a pseudorandom subset of $\B^n$, so
$\TV(Q, U)$ inherits the same bound up to a birthday-bound term
$O(\sum_x K(x) / 2^n)$.
\end{theorem}

\begin{proof}
Under the random oracle model, distinct $(x, k)$ pairs map to distinct
cipher values with high probability, so each $v \in \mathrm{im}(\enc)$
is the image of exactly one pair $(x, k)$ and $Q(v) = D(x)/K(x)$.  With
$K(x) = \lceil c\, D(x) \rceil$, rounding gives
\[
\frac{1}{c + 1/D(x)} \;\leq\; \frac{D(x)}{K(x)} \;\leq\; \frac{1}{c},
\]
so $Q(v) \in [1/(c + 1/\min_x D(x)),\ 1/c]$ for every $v$ in the image.
The image has $\sum_x K(x) \in [c,\ c + |X|]$ values, each receiving
mass approximately $1/c$, matching the uniform distribution on the
image to within the rounding error $|X|/(2\sum_x K(x))$.  The
image-to-$\B^n$ step follows from the random-oracle assumption: a
pseudorandom image of size $\sum_x K(x)$ inside $\B^n$ contributes at
most $O(\sum_x K(x) / 2^n)$ additional TV distance.
\end{proof}

\paragraph{Cost.}
Multiple representations cost space: each element now occupies $K(x)$
slots in the cipher map, so the total space is
$\sum_x K(x) \cdot (-\log_2 \varepsilon + H(Y))$ bits.  With
$K(x) = \lceil c \cdot D(x) \rceil$, the total number of
representations is $\sum_x K(x) \approx c$ (plus a rounding term bounded
by $|X|$).  The cost scales with the normalizing constant $c$, which
sets the achievable $\delta$ via Theorem~\ref{thm:multiplicity}.

\begin{example}[Homophonic encryption for Zipf-distributed queries]
\label{ex:homophonic}
A vocabulary of $m = 10{,}000$ words with Zipf distribution
$D(x_i) \propto 1/i$ has harmonic normalizer
$H_m = \sum_{i=1}^m 1/i \approx 9.788$, entropy
$H(D) \approx 9.55$ bits, and $H^* = \log_2 10{,}000 \approx 13.29$
bits, giving baseline $e \approx 0.72$.  Flattening the top
$b = 100$ words via $K(x_i) = \lceil c \cdot D(x_i) \rceil$ with
$c = 100 \cdot H_m \approx 979$ yields
$K(x_1) \approx 100, K(x_2) \approx 50, \ldots, K(x_{100}) \approx 1$,
total $\sum_{i=1}^{100} K(x_i) \approx 100 \cdot H_{100} \approx 519$
cipher cells for the top words.  Each of these cells has mass
$D(x_i)/K(x_i) \approx 1/979$, so the top-100 mass $\approx 0.53$ is
spread approximately uniformly across 519 cells.  The effective
entropy rises to $H(Q) \approx 11.5$ bits, giving $e \approx 0.87$.
The space overhead is approximately 519 additional trapdoors, a
$1.04\times$ increase over the baseline 10{,}000 trapdoors.
\end{example}

\subsection{Encoding Granularity}
\label{subsec:granularity}

Representation uniformity is a marginal property: it says the cipher
value distribution for each individual element is close to uniform.  It
says nothing about the joint distribution of multiple cipher
values.  The \emph{encoding granularity} controls what correlations are
hidden.

\begin{definition}[Entanglement parameter~{\cite[Sec.~9]{towell2026cipher}}]
For a system encoding $k$ correlated values, the \emph{entanglement
parameter} $p$ is the number of values encoded as a single cipher map
unit.  The system has type $\mathrm{cipher}(\B^p)^{k/p}$, where each
block of $p$ values is encoded jointly.
\end{definition}

\begin{proposition}[Granularity spectrum~{\cite[Prop.~9.1]{towell2026cipher}}]
\label{prop:granularity}
For a cipher map encoding a pair of values $(a, b) \in A \times B$:
\begin{enumerate}
\item \textbf{Joint encoding} ($p = 2$): the pair $(a, b)$ is encoded as
  a single cipher value.  The untrusted machine cannot distinguish
  $(a_1, b_1)$ from $(a_2, b_2)$ (up to $\delta$), but projections
  require cipher map constructions by the trusted machine.

\item \textbf{Component-wise encoding} ($p = 1$): $a$ and $b$ are
  encoded independently.  Projections are free, but the joint distribution
  of $(a, b)$ is preserved in the cipher pair.  With $N$ observations, the
  adversary estimates the joint distribution to accuracy $\xi$ in TV
  distance from $O(|A| \cdot |B| / \xi^2)$ samples.
\end{enumerate}
\end{proposition}

The encoding granularity creates a spectrum of confidentiality/
functionality trade-offs, analyzed in detail
in~\cite{towell2026algebraic}:

\begin{center}
\begin{tabular}{@{}lllll@{}}
\toprule
Granularity & Intermediates & Correlation hiding & Space & Projections \\
\midrule
Root ($p = k$) & 0 & Full & $O(|Y|^k)$ & None free \\
Intermediate & few & Partial & moderate & Some free \\
Leaf ($p = 1$) & $k$ & None (marginal only) & $O(k \cdot |Y|)$ & All free \\
\bottomrule
\end{tabular}
\end{center}

The sum-type impossibility theorem~\cite[Thm.~4.1]{towell2026algebraic}
shows that this trade-off is fundamental for sum types: for
$A + B$, tag hiding and untrusted pattern matching are mutually
exclusive.  Products leak correlations; sums leak the tag.

\paragraph{Confidentiality bound from orbit closure.}
The orbit closure~\cite[Sec.~5]{towell2026algebraic} quantifies the
dynamic confidentiality cost of exposing operations.  Given cipher maps
$F = \{\fhat_1, \ldots, \fhat_m\}$ available to the untrusted machine
and a starting cipher value $c$, the \emph{orbit}
$\orbitF_F(c)$ is the set of all cipher values reachable from $c$ by
composing operations in $F$.  The \emph{confidentiality}
$\mathrm{conf}_F(c)$ measures how much uncertainty the adversary retains
about the latent value encoded by $c$: it is $1$ when the adversary has
learned nothing and $0$ when the latent value is uniquely determined.
The orbit size bounds it~\cite[Thm.~5.3]{towell2026algebraic}:
\begin{equation}
\mathrm{conf}_F(c) \geq 1 - \frac{|\orbitF_F(c)|}{2^n}.
\end{equation}
A larger orbit (more reachable cipher values) means the adversary has
explored more of the cipher space, leaving fewer possibilities for the
latent value.  Fewer exposed operations means smaller orbit means higher
confidentiality.  This provides a formal justification for the root
encoding strategy: exposing no intermediate cipher values gives
$|\orbitF_\emptyset(c)| = 1$ and confidentiality $\geq 1 - 2^{-n}$.


%% ====================================================================
\section{Compositional Confidentiality}
\label{sec:composition}
%% ====================================================================

When cipher maps compose, as in $\ghat \circ \fhat$ or Boolean
combinations like $\widehat{\mathrm{AND}}(\fhat(c), \ghat(c))$,
confidentiality behaves differently from the single-query case.  Two
effects interact.  The primary effect, which motivates this section, is
\emph{correlation leakage}: marginal $\delta$-uniformity does not
prevent the adversary from recovering joint distributions when the
same cipher value appears in multiple evaluations.  The secondary
effect, inherited from~\cite{towell2026cipher}, is error compounding
through Boolean chains.  We treat correlation leakage first because
it exposes a fundamental limit of the constructions
in~\S\ref{sec:levers}: reducing $\delta$ to zero is necessary but not
sufficient when variables recur.

\subsection{Correlation Leakage from Shared Variables}
\label{subsec:correlation-leakage}

Even when each cipher map has uniform marginal output, evaluating multiple
cipher maps on the same cipher value leaks
correlations~\cite[Sec.~9.2]{towell2026cipher}.  If the untrusted machine
evaluates $\fhat_1(c)$ and $\fhat_2(c)$ for the same $c$, the pair
$(\fhat_1(c), \fhat_2(c))$ is a deterministic function of $c$, and the
joint distribution preserves the latent correlation between $f_1$ and
$f_2$.  This is the central compositional result of the paper.

\begin{theorem}[Compositional leakage]
\label{thm:comp-leakage}
Let $\fhat_1, \fhat_2$ be cipher maps for latent $f_1 : X \to Y_1$,
$f_2 : X \to Y_2$, each with $\delta_i \approx 0$ so their marginal
cipher value distributions are near-uniform on $\B^n$.  Assume the
untrusted machine observes pairs $(\fhat_1(c_i), \fhat_2(c_i))$ for
$i = 1, \ldots, N$, where each $c_i$ is an independent in-domain
cipher value drawn according to the pushforward of $D$ under $\enc$
(so the latent pair $(f_1(x), f_2(x))$ is drawn i.i.d.\ from the
true joint under $D$).  Then:
\begin{enumerate}
\item Each marginal cipher output is individually $\delta_i$-uniform on
  $\B^n$ (by hypothesis).
\item Mutual information is preserved:
  $I(\fhat_1(C); \fhat_2(C)) = I(f_1(X); f_2(X))$, where
  $X \sim D$ and $C = \enc(X, k)$ for uniform $k$.
\item The adversary estimates the joint distribution on
  $Y_1 \times Y_2$ to TV accuracy $\xi$ from
  $N = O(|Y_1| \cdot |Y_2| / \xi^2)$ samples, the standard plug-in
  estimator rate.
\item Any downstream cipher map $\fhat_3$ with Shannon-optimal
  acceptance predicate for the \emph{marginal} output distribution
  remains non-uniform when applied to the correlated pair:
  achieving $\delta_3 \approx 0$ requires reconstructing $\fhat_3$ with
  respect to the joint distribution.
\end{enumerate}
\end{theorem}

\begin{proof}
Part (1) is the hypothesis.
Part (2): the maps $\dec_i \circ \fhat_i$ push cipher pairs back to
latent pairs deterministically (up to $\eta$), so under the specified
sampling, the observed joint and the latent joint have equal mutual
information; data-processing gives equality because the push maps are
surjective onto $Y_1, Y_2$ on the in-domain part.
Part (3): the empirical distribution on a discrete support of size
$|Y_1| \cdot |Y_2|$ converges in TV to the true distribution at rate
$O(\sqrt{|Y_1|\cdot|Y_2|/N})$~\cite{cover2006elements}, so
$N = O(|Y_1|\cdot|Y_2|/\xi^2)$ samples suffice for TV accuracy $\xi$.
Part (4): when $I(f_1; f_2) > 0$, the joint input distribution to
$\fhat_3$ is not uniform on $Y_1 \times Y_2$, so a predicate optimized
for the product of marginals cannot achieve $\delta_3 = 0$; the
optimal $\delta_3$ is bounded below by the TV distance between the
joint and the product of marginals.
\end{proof}

\begin{remark}[Sampling regimes]
The bound in part (3) assumes each observation uses an independent
$c_i$ drawn from the same distribution.  A different regime, in which
the same cipher value $c$ is reused across evaluations, does not
yield additional samples of the joint and so has no corresponding
rate.  The interesting case for compositional leakage is many
distinct $c_i$ observed across deployment, each evaluated under both
$\fhat_1$ and $\fhat_2$.
\end{remark}

\paragraph{Why this matters.}
The constructions in~\S\ref{sec:levers} all target the marginal
parameter $\delta$.  Theorem~\ref{thm:comp-leakage} shows that
$\delta \to 0$ is insufficient whenever the untrusted machine
observes two evaluations on a shared cipher value: the joint is
recoverable at standard parametric rate.  Mitigating compositional
leakage requires either reducing observations ($N \lesssim |Y_1|\cdot
|Y_2| / \xi^2$), reducing granularity by joint encoding (Prop.~\ref{prop:granularity}), or injecting noise (\S\ref{subsec:noise}).

\paragraph{Mitigation strategies.}
Two approaches, both costly~\cite[Sec.~9.2]{towell2026cipher}:

\begin{description}
\item[Build the composition directly.]  Construct a single cipher map for
  $g(x) = f_3(f_1(x), f_2(x))$.  The untrusted machine sees only
  input and final output; no intermediates are exposed.  The domain is
  $|X|$ (same as either component), but the construction must evaluate
  $f_1$, $f_2$, and $f_3$ for every $x \in X$ at build time.

  A partial merge is also possible: merge $f_1$ and $f_2$ into a tuple
  cipher map $\widehat{(f_1, f_2)}(c) \to \cipher{Y_1 \times Y_2}$,
  then build $\fhat_3$ over the product codomain.
  This hides the correlation between $f_1$ and $f_2$ but the codomain
  $|Y_1| \times |Y_2|$ grows as a product of the component codomains.
  For $k$ merged components, the codomain is $\prod_i |Y_i|$, which
  grows exponentially.
  This is the fundamental space cost of the encoding granularity
  principle: hiding correlations requires encoding into product spaces.

\item[Compose components and inject noise.]  Keep practical
  component-wise evaluation and add $R$ noise queries per $N$ real
  queries.  Cost: bandwidth, not space.  The adversary's correlation
  estimates are diluted by the noise fraction.
\end{description}

\subsection{Error Compounding in Boolean Chains}

For Boolean-valued cipher maps (e.g., set indicator functions
$\hat{1}_A : \cipher{X} \to \cipher{\mathrm{Bool}}$, or any
$\fhat$ with $Y = \{\mathrm{True}, \mathrm{False}\}$), the
composition behavior depends on the gate type.  Let $p_T$ denote the
\emph{false positive rate} (FPR):
\[
p_T = \Pr[\dec(\fhat(\enc(x,k))) = \mathrm{True} \mid f(x) = \mathrm{False}],
\]
the probability that an element with $f(x) = \mathrm{False}$ is
decoded as True.
For in-domain elements with $\eta = 0$, FPR is zero by construction.
For out-of-domain elements (not in the construction domain), the hash
output is effectively random and lands in the True region with
probability $|T|/2^n$.

\begin{theorem}[FPR compounding~{\cite[Sec.~8]{towell2026cipher}}]
\label{thm:fpr-compounding}
For $k$ independent Boolean-valued cipher maps composed via Boolean
operations:
\begin{align}
\text{AND of } k \text{ tests:} &\quad
  \mathrm{FPR}_{\mathrm{total}} = p_T^k, \\
\text{OR of } k \text{ tests:} &\quad
  \mathrm{FPR}_{\mathrm{total}} = 1 - (1 - p_T)^k.
\end{align}
\end{theorem}

AND drives false positive rate down exponentially: each additional
conjunct is an independent filter.  OR drives it up: each additional
disjunct adds false positive mass.  NOT is approximate due to
complement non-preservation via the pigeonhole
principle~\cite{towell2026cipher}.

\paragraph{Effect on confidentiality.}
AND chains improve precision (fewer false positives) but maintain the
same recall (elements with $f(x) = \mathrm{True}$ pass all tests).
From a confidentiality perspective, AND chains reduce the effective noise
level: with $p_T = 0.05$ and $k = 3$, the FPR drops to
$0.05^3 \approx 0.000125$, meaning almost all decoded True outputs
correspond to elements where $f(x) = \mathrm{True}$.
The adversary can infer with higher confidence that a True result is
correct (lower entropy ratio $e$), a direct trade-off between precision
and confidentiality.

OR chains have the opposite effect: FPR increases, adding noise to the
output stream.
The adversary's uncertainty grows (higher $e$), but the system returns
more false positives, reducing its usefulness.

\paragraph{Convergence under deep composition.}
Long AND chains converge toward False (or noise).
With $p_T = 0.05$ and $p_F = 0.90$, each AND step multiplies the
probability of a True result by $p_T$ for noise inputs.
After $k$ ANDs with independent random cipher values, the probability
of a True output is at most $p_T^k$, which vanishes rapidly.
The cipher Boolean output after a deep AND chain is almost certainly
False or noise, regardless of the initial value.
This limits the depth of useful AND composition: beyond a few
terms, the signal (legitimate True results) is indistinguishable from
the False/noise floor.

Dually, long OR chains converge toward True (or noise).
Each OR step with an independent input produces True with probability
at least $p_T$, and after $k$ steps the probability of \emph{not} seeing
True is at most $(1 - p_T)^k$.
Deep OR chains saturate at True, again drowning the signal.

In both cases, noise acts as an attractor: AND pulls toward False/noise,
OR pulls toward True/noise.
The noise region prevents the adversary from distinguishing
``legitimately False after AND'' from ``noise-induced False,'' providing
a floor of uncertainty even in deep compositions.

\paragraph{Active probing via Boolean operations.}
Exposing AND and OR to the untrusted machine creates an orthogonal
confidentiality risk: the adversary can actively probe cipher values to
learn their latent meaning.
Given a cipher value $c$ and cipher AND, the adversary computes
$\mathrm{AND}(c, c)$.
Since cipher maps are deterministic, this always returns the same result
for a given $c$.
If the adversary also has cipher values $c_1, c_2$ and computes
$\mathrm{AND}(c_1, c_2)$, $\mathrm{AND}(c_1, c_1)$, and
$\mathrm{AND}(c_2, c_2)$, it can check whether the results are
structurally consistent with both being True, both False, or one of each.
Over many such probes, the adversary partitions cipher values into
equivalence classes that correspond to latent Boolean values.

This is exactly the orbit closure
attack~\cite[Sec.~5]{towell2026algebraic}: each Boolean operation
enlarges the orbit of reachable cipher values from $c$, and the
confidentiality degrades as
$\mathrm{conf}_F(c) \geq 1 - |\orbitF_F(c)|/2^n$.
The more operations the untrusted machine can apply, the larger the
orbit, the lower the confidentiality.
This creates a fundamental tension: exposing AND/OR/NOT enables
untrusted-side Boolean search (a functionality gain) but simultaneously
provides the adversary with tools to probe the cipher space
(a confidentiality cost).

One mitigation: \emph{typed composition chains}~\cite[Sec.~5.4]{towell2026algebraic}.
Instead of a single shared cipher Boolean space where AND can be
self-composed indefinitely, define a tower of distinct cipher spaces
$\cipher{\mathrm{Bool}}_0, \cipher{\mathrm{Bool}}_1, \ldots$
where $\mathrm{AND}_i$ maps from level $i$ to level $i+1$.
Without $\mathrm{AND}_{k}$, the chain terminates at depth $k$.
The orbit is bounded by $1 + k$ and the type system enforces the
limit at construction time.

%% ====================================================================
\section{Practical Measurement}
\label{sec:measurement}
%% ====================================================================

\subsection{Compression-Based Entropy Estimation}

The entropy of a sequence can be estimated without an explicit
probabilistic model by compressing it.  This follows from Shannon's
source coding theorem~\cite{shannon1948mathematical}: the expected length
of the output of an optimal lossless compressor equals the entropy of the
source.

\begin{proposition}[Compression estimator]
\label{prop:compression}
Let $\mathbf{c} = (c_1, \ldots, c_N)$ be a sequence of cipher values
observed by the untrusted machine, encoded as a bit string via some
fixed encoding $\mathrm{Encode}$.  Let $\mathrm{Compress}$ be a lossless
compressor.  Then:
\begin{equation}
\hat{H} = \frac{|\mathrm{Compress}(\mathrm{Encode}(\mathbf{c}))|}{N}
\end{equation}
is a positively biased estimator of the per-element entropy
$H(C_1, \ldots, C_N) / N$.  That is, $\hat{H} \geq H/N$ in expectation,
with equality when $\mathrm{Compress}$ is optimal.
\end{proposition}

\begin{proof}
An optimal compressor produces output whose expected bit length equals
$H(C_1, \ldots, C_N)$ by Shannon's source coding theorem.
Any sub-optimal compressor produces output at least as long.
The particular encoding is irrelevant: the compressor compensates for
the encoding overhead (as long as the encoding is uniquely decodable).
\end{proof}

The compression estimator has three practical advantages:
\begin{enumerate}
\item It requires no model of the query distribution.
\item It captures all forms of structure (frequency, correlation,
  temporal patterns) that a lossless compressor can detect.
\item It is cheap to compute: a single pass through the data with
  \texttt{gzip} or \texttt{zstd}.
\end{enumerate}

\paragraph{Entropy ratio estimator.}
Given the compression-based estimate $\hat{H}$ and a computable
$H^*$ (from the system parameters), the entropy ratio is estimated as:
\begin{equation}
\hat{e} = \frac{\hat{H}}{H^*}.
\end{equation}
Since $\hat{H}$ is positively biased, $\hat{e}$ is also positively
biased (overestimates confidentiality).  This is the conservative
direction: the system is at least as confidential as the estimate suggests.

\subsection{Monte Carlo Estimation}

When the query distribution is available (e.g., from historical logs),
entropy can be estimated by sampling.  Generate $M$ independent query
sequences of length $N$, compute the cipher value distribution, and
estimate $H$ directly:
\begin{equation}
\hat{H}_{\mathrm{MC}} = -\sum_v \hat{p}(v) \log_2 \hat{p}(v),
\end{equation}
where $\hat{p}(v)$ is the empirical frequency of cipher value $v$ across
the $M$ samples.  This estimator is negatively biased for finite $M$
(the Miller--Madow correction adds $(\hat{m} - 1) / (2M \ln 2)$ where
$\hat{m}$ is the number of distinct observed values), but converges to
the true entropy as $M \to \infty$.

\subsection{The Leakage Analyzer}

We define the \emph{leakage analyzer} as the procedure that combines
compression-based and Monte Carlo estimation:
\begin{enumerate}
\item Observe or simulate a sequence of cipher values.
\item Compute $\hat{H}$ via compression and $\hat{H}_{\mathrm{MC}}$ via
  sampling (if the distribution is available).
\item Compute $H^*$ from the system parameters using the maximum entropy
  formulas.
\item Report $\hat{e} = \hat{H} / H^*$ and the entropy gap
  $H^* - \hat{H}$ in bits.
\end{enumerate}
The entropy gap quantifies how many bits of information per query the
adversary potentially gains from the non-uniformity of the system.


%% ====================================================================
\section{Experimental Results}
\label{sec:experiments}
%% ====================================================================

We validate the theoretical predictions using the \texttt{cipher-maps}
Python library, which implements the batch cipher map construction with
perfect hash functions.  All experiments use the 20 Newsgroups corpus
(18,266 documents, 58,903 unique words).

\subsection{Boolean Search: Precision and Recall}

Cipher sets are built for each document using 8-bit cipher Booleans
($p_T = 0.05$, $p_F = 0.90$, $p_N = 0.05$).  Table~\ref{tab:boolean}
shows precision and recall for Boolean queries at 5,000 documents.

\begin{table}[ht]
\centering
\caption{Boolean search results on 20 Newsgroups (5,000 documents).}
\label{tab:boolean}
\begin{tabular}{@{}lcccc@{}}
\toprule
Query & Precision & Recall & FP & FN \\
\midrule
Single term & 0.39 & 1.00 & 248 & 0 \\
2-term AND & 0.76 & 1.00 & 12 & 0 \\
3-term AND & 0.97 & 1.00 & 1 & 0 \\
2-term OR & 0.25 & 0.97 & 310 & 8 \\
OR-AND-NOT & 0.31 & 0.88 & 285 & 15 \\
\bottomrule
\end{tabular}
\end{table}

The results confirm the FPR compounding theory:
AND drives false positives down ($248 \to 12 \to 1$), while OR and NOT
introduce noise.  From a confidentiality perspective, the single-term
query has the highest entropy ratio (most noise in the results), while
the 3-term AND has the lowest (almost deterministic results).

\subsection{FPR Compounding}

Table~\ref{tab:fpr} compares empirical and theoretical FPR for AND and
OR chains of length $k = 1$ through $5$, each with base FPR
$p_T = 0.05$.

\begin{table}[ht]
\centering
\caption{FPR compounding through Boolean chains ($p_T = 0.05$).}
\label{tab:fpr}
\begin{tabular}{@{}crrcrrc@{}}
\toprule
& \multicolumn{3}{c}{AND chain} & \multicolumn{3}{c}{OR chain} \\
\cmidrule(lr){2-4} \cmidrule(lr){5-7}
$k$ & Empirical & Theory & Ratio & Empirical & Theory & Ratio \\
\midrule
1 & 0.050 & 0.050 & 1.00 & 0.050 & 0.050 & 1.00 \\
2 & 0.003 & 0.0025 & 1.20 & 0.095 & 0.098 & 0.97 \\
3 & 0.000 & 0.000125 & --- & 0.142 & 0.143 & 0.99 \\
4 & 0.000 & 0.0000063 & --- & 0.186 & 0.185 & 1.01 \\
5 & 0.000 & 0.0000003 & --- & 0.224 & 0.226 & 0.99 \\
\bottomrule
\end{tabular}
\end{table}

The AND chain FPR drops below measurable levels by $k = 3$ (as
predicted by $0.05^3 \approx 10^{-4}$).  The OR chain FPR matches
$1 - (1 - 0.05)^k$ closely across all chain lengths.

\subsection{Encoding Granularity}

Table~\ref{tab:granularity} shows the cost spectrum for encoding
a 7-function decision pipeline (loan approval) at three granularity
levels, using a domain of 150 inputs (30 applicants $\times$ 5 loan
amounts).

\begin{table}[ht]
\centering
\caption{Encoding granularity spectrum for a 7-function decision pipeline.}
\label{tab:granularity}
\begin{tabular}{@{}lrrrr@{}}
\toprule
Granularity & Build (s) & Space (B) & Bits/elem & Exposed \\
\midrule
Root ($p = 7$) & 0.02 & 628 & 33.5 & 0 \\
Intermediate ($p \approx 3$) & 0.05 & 1,204 & 64.2 & 3 \\
Leaf ($p = 1$) & 0.08 & 2,116 & 112.9 & 7 \\
\bottomrule
\end{tabular}
\end{table}

The root encoding exposes zero intermediate values (maximum
confidentiality) at 33.5 bits per element.  The leaf encoding exposes all
7 intermediate cipher values (zero correlation hiding) at 112.9 bits per
element---3.4$\times$ the root cost.  The intermediate level balances
the trade-off: 3 exposed values, 64.2 bits per element.

\subsection{Case Study: Confidentiality Improvement}

We demonstrate the three levers on a system with vocabulary $m = 10{,}000$
and Zipf-distributed queries.

\begin{table}[ht]
\centering
\caption{Confidentiality improvement via the two constructions of
\S\ref{sec:levers}.  Values are analytical, computed from the Zipf
entropy, the mixture-entropy formula in Theorem~\ref{thm:noise-dilution},
and the homophonic construction in Example~\ref{ex:homophonic}.}
\label{tab:case-study}
\begin{tabular}{@{}lrrr@{}}
\toprule
Configuration & $e$ & Space overhead & Bandwidth overhead \\
\midrule
Baseline (simple substitution) & 0.72 & $1.00\times$ & $1.00\times$ \\
+ Homophonic ($b = 100$, $c \approx 979$) & 0.87 & $1.04\times$ & $1.00\times$ \\
+ Noise injection ($R/N = 0.5$) & 0.88 & $1.00\times$ & $1.50\times$ \\
Combined & 0.98 & $1.04\times$ & $1.50\times$ \\
Theoretical maximum & 1.00 & $\to \infty$ & $\to \infty$ \\
\bottomrule
\end{tabular}
\end{table}

The combined strategy improves $e$ from 0.72 to 0.98: a 4\% space
overhead and 50\% bandwidth overhead yields a 26 percentage point
confidentiality improvement.  The small space cost (compared to prior
analyses that assumed $K(x) \propto 1/D(x)$, which requires
$\sum_x K(x) \approx c \cdot m \cdot H_m$ slots) reflects the classical
homophonic prescription $K(x) \propto D(x)$: concentrating multiplicity
on the heavy part of $D$ is exponentially cheaper than spreading it
over the tail.


%% ====================================================================
\section{Discussion and Open Questions}
\label{sec:discussion}
%% ====================================================================

\paragraph{The entropy ratio as a design tool.}
The entropy ratio $e$ gives system designers a single number to
optimize.  Given a resource budget (space $S$, bandwidth $B$), the
designer solves: maximize $e$ subject to $\text{space} \leq S$ and
$\text{bandwidth} \leq B$.  The two constructions analyzed here (noise
injection, multiplicity) plus the inherited granularity dimension
provide the degrees of freedom.  Noise costs bandwidth but not space;
multiplicity costs space but not bandwidth; granularity trades
confidentiality for functionality.

\paragraph{Relationship to the orbit closure.}
The entropy ratio measures \emph{passive} confidentiality: what the
adversary learns by observing the cipher value stream.  The orbit closure
from~\cite{towell2026algebraic} measures \emph{active} confidentiality:
what the adversary learns by applying the cipher maps it holds.  A
complete confidentiality analysis requires both: high $e$ (passive) and
small orbit (active).  Connecting these two measures---showing that high
$e$ implies small effective orbit under certain conditions---is an open
problem.

\paragraph{Adaptive $K(x)$.}
The multiplicity $K(x)$ is set at construction time based on an assumed
distribution $D$.  If $D$ changes over time (e.g., query distribution
shift), the system loses uniformity.  Adaptive $K(x)$ that adjusts to
observed frequencies without reconstruction would improve robustness.
This requires an online construction strategy where representations can
be added incrementally.

\paragraph{Tight bounds for composition.}
The general composition theorem gives
$\eta_{g \circ f} \leq 1 - (1 - \eta_f)(1 - \eta_g)$ under
independence.  For specific circuits (e.g., 3-SAT evaluation over cipher
Booleans), the actual error rate depends on the input distribution and
the circuit structure.  Case-by-case interval arithmetic gives tighter
bounds but does not scale.  Automated analysis tools that propagate
entropy through circuit DAGs would be useful.

\paragraph{Beyond Boolean search.}
This paper focuses on encrypted Boolean search as the primary
application.  The cipher map framework supports arbitrary functions
$f : X \to Y$, and the confidentiality theory extends naturally.  Ranked
retrieval (scoring functions), phrase search (bigram models), and
approximate string matching all admit cipher map implementations with
different confidentiality profiles.

\paragraph{Limitations.}
\begin{enumerate}
\item The entropy ratio is an average measure; it does not bound the
  leakage of any \emph{specific} query.  A single high-frequency query
  may be identifiable even when $e$ is close to 1.

\item Compression-based estimation is positively biased (overestimates
  confidentiality).  The bias decreases with sequence length but can be
  significant for short sequences.

\item The analysis assumes the random oracle model for cryptographic
  hashing.  In practice, hash functions have biases that may reduce
  confidentiality below the theoretical predictions.

\item Marginal uniformity ($\delta \approx 0$) is necessary but not
  sufficient.  Joint distributions, temporal patterns, and side channels
  (timing, bandwidth) can leak information that the entropy ratio does
  not capture.
\end{enumerate}


%% ====================================================================
\section{Conclusion}
\label{sec:conclusion}
%% ====================================================================

We have developed a quantitative confidentiality theory for cipher map
systems, grounded in the cipher map framework
of~\cite{towell2026cipher}.  The entropy ratio $e = H/H^*$, borrowed
from the QIF literature, becomes operational in this setting via the
Fannes bridge: the representation-uniformity parameter $\delta$
lower-bounds $e$ through the Fannes--Audenaert continuity inequality,
reducing the design problem to minimizing $\delta$.  Two constructions
reduce $\delta$ with explicit costs---noise injection (bandwidth) and
multiple representations with $K(x) \propto D(x)$ (space).  A third
dimension, encoding granularity, is inherited from the cipher maps
framework.  The three together are expressible in terms of the cipher
map parameters $(\eta, \varepsilon, \delta, \mu)$ and the entanglement
parameter $p$.

The key insight is that the cipher map's four properties already
determine the confidentiality profile.  Totality enables noise injection
(filler queries are indistinguishable from real queries).  Representation
uniformity enables frequency hiding (multiple representations flatten the
distribution).  Composability enables chained evaluation but introduces
correlation leakage.  Correctness provides plausible deniability
($\eta > 0$ means the adversary cannot be certain of any decoded output).

Experimental validation on the 20 Newsgroups corpus confirms the
theoretical predictions: FPR compounds as $p_T^k$ for AND chains and
$1 - (1 - p_T)^k$ for OR chains; the encoding granularity spectrum runs
from 33.5 bits/element (root, zero intermediates) to 112.9 bits/element
(leaf, all intermediates exposed); and the combined homophonic/noise
strategy improves confidentiality from 72\% to 98\% with approximately
$1.04\times$ space and $1.5\times$ bandwidth overhead.

The theory connects two previously separate lines of work: the cipher map
formalism (properties and composition) and the entropy-based
confidentiality analysis (ratios and maximum entropy).  Together, they
provide a complete framework for designing, measuring, and improving
confidentiality in systems that compute on data hidden behind a trapdoor.


\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
